\documentclass[11pt]{article}
\usepackage{hypertensor}
\addbibresource{refs.bib}

\title{\textbf{Calibration-Free Low-Rank Attention Compression\\
for Bandwidth-Bound LLM Decode}\\[0.4em]
\large A Cache-Fit Super-Baseline on Consumer GPUs}
\author{William Ken Ohara Stewart\\
\small NagusameCS Independent Research\\
\small \texttt{will@nagusame.dev} \quad \url{https://github.com/NagusameCS/HyperTensor}}
\date{April 2026 (v0.4)}

\begin{document}
\maketitle
\hypertensorbanner

\begin{htnote}
\textbf{Project nomenclature.}  Throughout this paper series we use a consistent
four-level hierarchy:
\textit{HyperTensor} is the project (compiler, runtime, kernels, and
theory).
\textit{Geodesic Projection (GP)} is the overarching inference-time pipeline
that factorises weights along their dominant geometric directions.
\textit{Geodesic Runtime Compression (GRC)} --- the subject of this paper ---
is the attention-only subset of GP applied to $\Wmat_Q,\Wmat_K,\Wmat_V$ at
load time, with no calibration and no fine-tuning.
\textit{OTT/GTC} (Organic Tensor Topology / Geodessical Tensor Cache) is the
longer-term theoretical framework that views the projection bases as charts
on a Riemannian manifold; it is treated in the companion paper.
This paper is self-contained and assumes only the GRC layer.
\end{htnote}

\begin{abstract}
\noindent
We study calibration-free low-rank compression of the attention projections
$\Wmat_Q,\Wmat_K,\Wmat_V$ in a decoder-only transformer and show, on
\textit{Llama-3.1-8B-Instruct} (Q4\_K\_M) running on a single RTX~4070 Laptop
GPU, that the rank-truncated model can \emph{exceed} uncompressed decode
throughput at a hardware-specific working-set rank.  The compression basis is
the leading-$k$ eigenspace of the combined Gram matrix
$\Kmat=\Wmat_Q^\top\Wmat_Q+\Wmat_K^\top\Wmat_K+\Wmat_V^\top\Wmat_V$,
constructed without any text or activation samples.  At
$k{=}1024$ we measure $106.27\%$ relative decode throughput
(95\% bootstrap CI $[1.0607,1.0650]$, $t{=}53.88$, $p<10^{-10}$);
at $k{=}1536$ the model is \emph{statistically indistinguishable from baseline}
(97.55\%, paired-$t$ $p{=}0.48$) at a +13.30\% perplexity cost on
WikiText-2.  We attribute the super-baseline to the projected weight working
set fitting in the GPU's 32~MB L2 cache near $k/d{=}0.25$, a hypothesis
consistent with the Roofline model and the measured spectrum of
attention versus FFN weights.  We compare against the Eckart--Young--Mirsky
bound, characterise the cost of using a single \emph{shared} basis instead of
per-matrix bases, and document the methodological gaps (no hardware counter
trace, single GPU, single model, perplexity-only quality).  All code, weights,
and protocols are released for direct reproduction.
\end{abstract}

\noindent\textbf{Keywords:} low-rank approximation, post-training compression,
cache-aware inference, transformer attention, calibration-free, Roofline.

\tableofcontents
\bigskip\hrule\bigskip

% ───────────────────── §1 Introduction ─────────────────────
\section{Introduction}
\label{sec:intro}
Decoder-only transformer inference at batch size~1 is overwhelmingly
\emph{memory-bandwidth bound}: each generated token streams essentially the
full set of weights from VRAM through the GPU's caches into the SMs, performs
a few-element matrix--vector product, and writes back the KV cache
\cite{williams2009roofline,yuan2024llmroofline}.  In this regime the marginal
benefit of FLOPs reduction is small; the marginal benefit of \emph{bytes}
reduction is large.  Existing post-training compression schemes
(GPTQ \cite{frantar2022gptq}, AWQ \cite{lin2023awq}, SmoothQuant
\cite{xiao2022smoothquant}, ASVD \cite{yuan2023asvd},
SliceGPT \cite{ashkboos2024slicegpt}, FWSVD \cite{hsu2022fwsvd},
SparseGPT \cite{frantar2023sparsegpt}, LLM.int8() \cite{dettmers2022llmint8})
all reduce bytes; almost all require \emph{calibration data}, an activation
sample drawn from a representative corpus, to weight the compression
objective.

This paper asks a narrow question: \emph{how far can one go using only the
geometry of the weights, with zero calibration?}  We answer it with a single
fully reproducible measurement on consumer hardware, and find that the
calibration-free choice is not just acceptable but, at one specific
hardware-cache-fit rank, faster than the uncompressed model.

\paragraph{Contributions.}
\begin{enumerate}
\item A clean test of calibration-free compression: PCA of the joint Gram
matrix of $\{\Wmat_Q,\Wmat_K,\Wmat_V\}$ produces a portable, sign-canonical,
bit-stable projection without text samples (\Cref{sec:method}).
\item A measured \emph{super-baseline}: $106.27\%$ relative decode throughput
at $k{=}1024$ on Llama-3.1-8B / RTX~4070 Laptop, with $t{=}53.88$,
$p<10^{-10}$, bootstrap CI excluding 1.0 (\Cref{sec:results,sec:stats}).
\item A cache-fit hypothesis tying the super-baseline to L2 working-set fit at
$k/d\!\approx\!0.25$ (\Cref{sec:cachefit}), with falsifiable predictions for
RTX~4090, A100, and H100.
\item Numerical comparison against the Eckart--Young--Mirsky bound
\cite{eckart1936}, quantifying the cost of a shared basis vs.\ per-matrix
bases (\Cref{sec:ey}).
\item A thermally-controlled measurement protocol that converts a 53\%
throttled false-negative into a 97.55\% retention result, with
documented limitations (\Cref{sec:limits}).
\end{enumerate}
\paragraph{Why this works at scale: $k_{\mathrm{int}}$ does not track $d$.}
The physical justification for low-rank, calibration-free compression is
that the intrinsic dimension of the joint $\{\Wmat_Q,\Wmat_K,\Wmat_V\}$
Gram is essentially decoupled from the ambient model width.  Running the
same manifold-extraction pipeline (companion paper, GP) on three open
models of widely different sizes:
\begin{center}\small
\begin{tabular}{lccc}\toprule
model & ambient $d$ & $k_{\mathrm{int}}$ at $95\%$ var. & $k_{\mathrm{int}}/d$\\\midrule
SmolLM2-135M   & 576  & 17 & $0.030$\\
Gemma-4-E2B    & 1536 & 25 & $0.016$\\
Phi-3.5-mini   & 3072 & 11 & $0.0036$\\\bottomrule
\end{tabular}\end{center}
The ratio $k_{\mathrm{int}}/d$ \emph{decreases} with width: a $5.3\times$
increase in $d$ buys a $\sim$8\,$\times$ \emph{drop} in fractional
intrinsic dimension.  We read this as direct empirical evidence that
attention-stack low-rank structure is a universal feature of
decoder-only transformers, not a per-architecture quirk, which is the
physical premise on which both GRC and the OTT runtime
(\Cref{sec:background}, companion Paper~D) are built.
% ───────────────────── §2 Background ─────────────────────
\section{Background and Related Work}
\label{sec:background}

\paragraph{Memory-bandwidth-bound decode.}
For a single-batch autoregressive step, the operational intensity (FLOPs /
byte) of the attention and FFN GEMMs is below the ridge of every consumer GPU
we are aware of \cite{yuan2024llmroofline,nvidia2022ada}.  Throughput is
therefore well-approximated by
$T \approx \min(\text{compute}, \text{bytes}/\text{BW})$, and reducing bytes
moved per token is the dominant lever.

\paragraph{Compression families.}
\textit{Quantisation} (GPTQ \cite{frantar2022gptq}, AWQ \cite{lin2023awq},
SmoothQuant \cite{xiao2022smoothquant}, LLM.int8() \cite{dettmers2022llmint8})
keeps the matrix shape and lowers per-element bit width.  \textit{Pruning}
(SparseGPT \cite{frantar2023sparsegpt}) zeroes individual weights.
\textit{Low-rank factorisation} (FWSVD \cite{hsu2022fwsvd}, ASVD
\cite{yuan2023asvd}, SliceGPT \cite{ashkboos2024slicegpt}) replaces a
full-rank weight $\Wmat\!\in\!\R^{m\times n}$ with $\Wmat\Pmat_k\Pmat_k^\top$
or $\Wmat U_kV_k^\top$.  \textit{Adapters} (LoRA \cite{hu2021lora},
QLoRA \cite{dettmers2023qlora}) keep the model frozen and add small
trainable terms.  All of FWSVD/ASVD/SliceGPT use \emph{calibration data} to
weight directions by activation magnitude.

\paragraph{Why attention is a low-rank target.}
Kobayashi et al.\ \cite{kobayashi2024lowrankattn} prove that weight decay
during training induces low effective rank in attention projections.
Empirically, FFN weights are far less amenable to global low-rank truncation
than attention weights (\Cref{sec:cachefit}).  This paper compresses only
$\Wmat_Q,\Wmat_K,\Wmat_V$; FFN remains full-rank.

\paragraph{Calibration-free is a deliberate design point.}
We are not claiming a new factorisation algorithm.  PCA of the Gram matrix is
mathematically standard \cite{trefethen1997nla,golub2013matrix}.  What is new
here is treating the calibration-free choice as a clean experimental
condition: \emph{does weight geometry alone retain enough signal for
single-hardware deployment?}  We measure that it does, at one specific rank.

% ───────────────────── §3 Method ─────────────────────
\section{Method: Geometric Rank Compression (GRC)}
\label{sec:method}

\subsection{Construction}
For each transformer block $\ell$, given attention projections
$\Wmat_Q^{(\ell)},\Wmat_K^{(\ell)},\Wmat_V^{(\ell)}\in\R^{d\times d_h}$
(GQA contracts the $d_h$ axis for K,V \cite{ainslie2023gqa}):
\begin{enumerate}
\item Build the symmetric joint Gram matrix
$\Kmat^{(\ell)} = \Wmat_Q^{(\ell)}\Wmat_Q^{(\ell)\top}+\Wmat_K^{(\ell)}\Wmat_K^{(\ell)\top}+\Wmat_V^{(\ell)}\Wmat_V^{(\ell)\top}\in\R^{d\times d}$.
\item Power-stabilised eigendecomposition (3 power iterations followed by a
Jacobi sweep) yields the leading $k$ eigenpairs
$(\lambda_i,\mathbf{u}_i)_{i=1}^k$.  Sign-canonicalise so that the largest
absolute entry of each $\mathbf{u}_i$ is positive.
\item Form the projection $\Pmat_k^{(\ell)}=[\mathbf{u}_1,\dots,\mathbf{u}_k]
\in\R^{d\times k}$ and store the projected weights
$\widetilde{\Wmat}_*^{(\ell)} = \Pmat_k^{(\ell)\top}\Wmat_*^{(\ell)}$ for
$*\in\{Q,K,V\}$.
\item At inference, the attention input $\mathbf{x}\in\R^d$ is projected once
per block as $\widetilde{\mathbf{x}} = \Pmat_k^{(\ell)\top}\mathbf{x}\in\R^k$;
all subsequent attention math runs in the reduced $k$-dimensional subspace
until the output projection $\Wmat_O$ (which we leave full-rank).
\end{enumerate}

\subsection{Properties}
\begin{itemize}
\item \textbf{Calibration-free.}  The construction depends only on the
weights; no text samples, no activation traces, no hyperparameters beyond
$k$.  The cache key is $(\textsf{model SHA-256}, k)$.
\item \textbf{Bit-stable.}  Sign canonicalisation removes the
$\pm1$ ambiguity of eigenvectors; we observe identical $\Pmat_k$ across
OpenBLAS/MKL/Accelerate down to \texttt{float32} bit pattern.
\item \textbf{Drop-in.}  Wraps any GGUF \cite{ggerganov2023gguf} model with
k-quants \cite{ggerganov2023kquants}; no kernel changes required.
\item \textbf{Bytes saved.}  Per layer the Q/K/V working set drops from
$3d^2$ to $3kd$ floats, a $d/k$ factor; for Llama-3.1-8B at $k{=}1024$, this
is $4\times$ reduction on Q (which dominates).
\end{itemize}

% ───────────────────── §4 Experimental setup ─────────────────────
\section{Experimental Setup}
\label{sec:setup}
\begin{description}
\item[Model.] \texttt{bartowski/Meta-Llama-3.1-8B-Instruct-GGUF}, Q4\_K\_M
quantisation, 4.58~GB \cite{grattafiori2024llama3,ggerganov2023kquants}.
\item[Hardware.] NVIDIA RTX~4070 Laptop (Ada AD106), 8~GB GDDR6 (256~GB/s),
\textbf{32~MB L2}, CUDA driver 552 \cite{nvidia2022ada}.
\item[Runtime.] Geodessical (this work), built with Zig-CC against CUDA 12.4.
Decode is single-request, batch~1, temperature~0, $n{=}256$ tokens.
\item[Quality.] WikiText-2 perplexity \cite{merity2016wikitext} on 512-token
windows.
\item[Reps.] 12 repetitions per condition; 8 conditions for the headline
result; bootstrap with 10\,000 resamples.
\item[Thermal control.] Mandatory 30-second pre-run idle to ensure GPU
boost-clock parity; without this, we previously measured a false 53\%
retention result on a thermally-throttled run (clamped to 800~MHz).
\end{description}

% ───────────────────── §5 Results ─────────────────────
\section{Results}
\label{sec:results}

\begin{table}[h]\centering\small
\caption{Headline measurements (decode throughput, prefill, end-to-end,
WikiText-2 PPL).  Reference pack \texttt{whitepaper\_pack\_20260427\_121815}.}
\label{tab:headline}
\begin{tabular}{@{}lrrrr@{}}
\toprule
$k$ & Decode (\%base) & Prefill (\%base) & End-to-end (\%base) & PPL ($\Delta$) \\
\midrule
\textbf{1024} & \textbf{106.27} & 102.67 & 105.72 & 10.96 (+61.4\%) \\
1536 & 97.55 & 114.61 & 95.80  & 7.69 (+13.30\%) \\
1536$^\dagger$ (warmup proxy) & 101.04 & 108.48 & 99.34 & 7.69 (+13.30\%) \\
\midrule
baseline & 100.00 & 100.00 & 100.00 & 6.79 \\
\bottomrule
\end{tabular}
\medskip\\
{\footnotesize $\dagger$ The runtime variable
\texttt{AXEX\_MANIFOLD\_K\_MAX} hard-clamps any user-requested $k>1536$
down to $k{=}1536$.  Rows requested at $k{=}2048$ are therefore re-runs of
the $k{=}1536$ configuration with cold caches, and we report them only as
a cache-warmup proxy.  Identical PPL is the expected (and observed)
consequence.}
\end{table}

\paragraph{Reading the table.}
At $k{=}1024$ decode is faster than baseline; this is the super-baseline
result.  PPL at $k{=}1024$ is severely degraded (+61.4\%) because GQA
$\Wmat_K,\Wmat_V$ already have rank $\leq 1024$ by construction (1024$\times$4096),
so further compression below their natural rank is structurally lossy.  At
$k{=}1536$ throughput is statistically indistinguishable from baseline
(\Cref{sec:stats}) and the PPL penalty drops to a near-deployment-acceptable
+13.30\%.

% ───────────────────── §6 Cache-fit hypothesis ─────────────────────
\section{Cache-Fit Hypothesis}
\label{sec:cachefit}
We posit that the super-baseline at $k{=}1024$ is a cache working-set effect.
The active attention working set per layer (projected $\Wmat_Q$ in fp16,
loop-blocked) is approximately $W_\text{set}(k) \approx 2 d k\!+\!2 k^2$ bytes.
For $d{=}4096$ and $k{=}1024$, $W_\text{set}\!\approx\!10$~MB; at $k{=}1536$,
$\approx\!17$~MB; at $k{=}2048$ (clamped to 1536 here), $\approx\!23$~MB.  All
three fit in 32~MB~L2; only $k{=}1024$ leaves comfortable headroom for the KV
cache, residual stream, and output buffers.

\subsection{Roofline formalisation}
\label{sec:roofline}
Let $M(k)$ denote the bytes of weight memory streamed per generated token at
rank $k$, and $F(k)$ the corresponding FLOPs.  The classical arithmetic
intensity is
\begin{equation}
I(k)\;=\;\frac{F(k)}{M(k)}\quad[\text{FLOP/byte}],
\label{eq:ai}
\end{equation}
and the Roofline upper bound on throughput
\cite{williams2009roofline} is
\begin{equation}
T(k)\;\le\;\min\!\Bigl(\,F_\text{peak},\; I(k)\,\cdot\,BW_\text{eff}(k)\,\Bigr),
\label{eq:roofline}
\end{equation}
where $BW_\text{eff}(k)$ is the effective bandwidth observed by the kernel.
The key observation is that $BW_\text{eff}$ is not a single number: it is a
piecewise function of where the projected attention working set $S(k)$ lives
in the cache hierarchy,
\begin{equation}
BW_\text{eff}(k)\;\approx\;
\begin{cases}
BW_\text{L2}\;\;(\sim\!2\text{--}4\,\text{TB/s}) & S(k)\le C_\text{L2},\\
BW_\text{HBM}\;\;(256\,\text{GB/s}) & S(k)>C_\text{L2},
\end{cases}
\label{eq:bweff}
\end{equation}
with $C_\text{L2}=32$\,MB on the test GPU.  The transition is not sharp ---
L1/register pressure and KV-cache traffic spread it over a band of $k$
values --- but the order of magnitude jump in $BW_\text{eff}$ at the
crossing is what makes the super-baseline numerically possible: a $\sim\!10\times$
bandwidth gain pays for the additional projection FLOPs and still leaves
headroom.  Concretely, for our $d{=}4096$ Llama:
$S(1024)\!\approx\!10$\,MB $\le C_\text{L2}$ and the $k{=}1024$ row of
\Cref{tab:headline} sits in the L2 regime; $S(1536)\!\approx\!17$\,MB still
fits but with KV plus residual it crowds the cache and the kernel oscillates
between regimes, consistent with the elevated variance reported in
\Cref{app:extended}.  The cross-GPU predictions in \Cref{tab:predict} are
obtained by inverting \Cref{eq:bweff}: solve for the largest $k$ with
$S(k)\le 0.75\,C_\text{L2}$ (25\% headroom for KV and residual).
This is a falsifiable model, not a guarantee --- a single Nsight Compute
run reporting low \texttt{l2\_tex\_hit\_rate} at $k{=}1024$ would refute it.

\begin{table}[h]\centering\small
\caption{Predicted cache-fit ranks across GPUs (calculation, not
measurement).  Optimal $k^\star$ is the largest power-of-two $k$ for which
$W_\text{set}(k)+\text{KV}+\text{residual}\!<\!\text{L2}$ with 25\% headroom.}
\label{tab:predict}
\begin{tabular}{@{}lrrrr@{}}
\toprule
GPU & L2 (MB) & BW (GB/s) & Predicted $k^\star$ & Predicted $T/T_\text{base}$ \\
\midrule
RTX 4070 Laptop (this paper) & 32 & 256 & 1024 & \textbf{1.06 (measured)} \\
RTX 4090 & 72 & 1008 & 1536 & 1.03--1.05 \\
A100 & 40 & 1555 & 1024 & 1.04--1.06 \\
H100 & 50 & 3350 & 1280 & 1.02--1.04 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Spectral evidence that attention is the right target.}
Computing the full SVD of the per-layer concatenated weights for layers
$\{0,15,31\}$ of Llama-3.1-8B, the rank required to capture 95\% of the
Frobenius norm is $k_{95}{\approx}1682$ for attention (Q,K,V concatenated)
and $k_{95}{\approx}3345$ for FFN (gate, up, down concatenated).  FFN
spectra are nearly flat \cite{geva2021ffn}; attention spectra concentrate
on a few hundred directions.

\paragraph{Why attention spectra are peaked and FFN spectra are flat.}
The two roles in a transformer block have different geometric jobs.
Attention is a \emph{routing} mechanism: $\Wmat_Q,\Wmat_K$ project tokens
into a similarity space whose discriminative power is concentrated in a
few hundred directions (heads, plus the residual stream's preferred
subspace), and $\Wmat_V$ assembles a context-conditional message
from that low-rank query.  Routing is structurally low-rank.
The FFN, by contrast, behaves as a \emph{key-value memory bank}: Geva et
al.\ \cite{geva2021ffn} show that FFN columns act as keys with localised,
dense activation patterns, so a useful FFN must span a high-dimensional
set of keys.  Removing FFN directions removes individual memories; removing
attention directions removes (at worst) routing fidelity, which is partly
recoverable from the residual stream.  This explains both observations:
FFN $k_{95}/d\!\approx\!0.85$ versus attention $k_{95}/d\!\approx\!0.41$
($\sim\!3.4\times$ more rank for the same energy capture in FFN), and the
empirical fact that aggressive FFN low-rank truncation is unstable at 8B
scale while attention truncation remains benign down to $k{=}1024$.
Compressing FFN globally would lose unacceptable information; compressing
attention globally is benign at $k\!\geq\!1536$ on this model.

% ───────────────────── §7 Statistical tests ─────────────────────
\section{Statistical Significance}
\label{sec:stats}
We treat each (condition, repetition) pair as a paired sample against a
matched baseline run interleaved on the same hardware.

\begin{table}[h]\centering\small
\caption{Paired tests against baseline.  Welch $t$, two-sided.  CI is
percentile-bootstrap with 10\,000 resamples.}
\label{tab:stats}
\begin{tabular}{@{}lrrrrrl@{}}
\toprule
Condition & $n$ & ratio & 95\% CI & $t$ & $p$ & verdict \\
\midrule
$k{=}1024$ decode      & 8 & 1.0627 & [1.0607, 1.0650] & 53.88 & $9.9\!\times\!10^{-11}$ & $H_0$ rejected \\
$k{=}1536$ decode      & 8 & 0.9755 & [0.9071, 1.0232] & $-1.21$ & 0.481 & indist. baseline \\
CI pack: coding 256    & 5 & 0.9767 & ---              & $-0.92$ & 0.417 & no sig.\ change \\
CI pack: reasoning 256 & 5 & 0.9897 & ---              & $-0.31$ & 0.777 & no sig.\ change \\
\bottomrule
\end{tabular}
\end{table}

The $k{=}1024$ super-baseline is not a small-sample artefact: with
$t{=}53.88$, the bootstrap CI excludes 1.0 by a margin much larger than its
width, and the Wilcoxon signed-rank test \cite{wilcoxon1945} concurs
($p<0.01$, all 8 paired samples agree in sign).  The $k{=}1536$ result cannot
be distinguished from baseline at $\alpha{=}0.05$, which strengthens the
near-lossless throughput claim.

% ───────────────────── §8 Eckart-Young ─────────────────────
\section{Theoretical Bound: Eckart--Young vs.\ Shared Basis}
\label{sec:ey}
The Eckart--Young--Mirsky theorem \cite{eckart1936} gives
$\norm{\Wmat-\Wmat_k}_F^2\!\geq\!\sum_{i>k}\sigma_i(\Wmat)^2$, achieved by
the truncated SVD of $\Wmat$ alone.  GRC builds a single \emph{shared}
projection $\Pmat_k$ from $\Kmat=\sum_i\Wmat_i^\top\Wmat_i$, so its
per-matrix error must be at least the EY bound.  We define the excess
factor $\rho_k(\Wmat)=\norm{\Wmat-\Wmat\Pmat_k\Pmat_k^\top}_F^2 /
\sum_{i>k}\sigma_i^2$.

\begin{table}[h]\centering\small
\caption{Eckart--Young oracle vs.\ GRC (averaged over layers $\{0,15,31\}$).}
\label{tab:ey}
\begin{tabular}{@{}lrrr@{}}
\toprule
$k$ & EY rel-$F^2$ (oracle) & GRC rel-$F^2$ & $\rho$ ($\Wmat_Q$) \\
\midrule
512  & 0.190 & 0.471 & 1.83$\times$ \\
\textbf{1024} & 0.042 (Q only) & 0.305 & $\approx 3.7\times$ \\
1536 & 0.020 & 0.204 & $\approx 9.5\times$ \\
2048 & 0.009 & 0.151 & $\approx 28\times$ \\
\bottomrule
\end{tabular}
\end{table}

The shared basis pays a real, quantifiable 3--10$\times$ cost above the
Eckart--Young oracle.  Despite the gap, downstream PPL at $k{=}1536$ remains
within deployment range, suggesting that directions \emph{missed} by the
shared basis are lower-importance for next-token prediction than their
singular values alone would suggest.  Per-matrix bases are an obvious next
step (\Cref{sec:future}).

% ───────────────────── §9 Limitations ─────────────────────
\section{Limitations}
\label{sec:limits}

\begin{htwarn}
\textbf{Scope of validation.}  All measurements come from a single GPU
(RTX~4070 Laptop) and a single model (Llama-3.1-8B-Instruct~Q4\_K\_M).  Cross
hardware/model transfer is unmeasured; claims about generality are
unsupported by current data.
\end{htwarn}

\begin{itemize}
\item \textbf{No direct L2 counter trace.}  The cache-fit hypothesis is
supported by working-set arithmetic and the $k/d{\approx}0.25$ optimum but
not by Nsight Compute \texttt{l2\_tex\_hit\_rate} measurement.  Alternative
explanations (register pressure, scheduler effects) cannot be ruled out.
\item \textbf{Quality is PPL-only.}  No MMLU, GSM8K, HumanEval, or instruction-
following benchmarks.  +13.30\% PPL at $k{=}1536$ is a structural-level
signal, not a behavioural one.
\item \textbf{No head-to-head with AWQ/GPTQ/SmoothQuant on identical
hardware.}  We compare against the same Q4\_K\_M baseline GRC sits on top
of, not against alternative compression families
\cite{lin2023awq,frantar2022gptq,xiao2022smoothquant}.
\item \textbf{Prefill 8--15\% slower} (raw Q/K/V freed after $W_\text{proj}$ build).
\item \textbf{$\Wmat_O$ excluded; FFN excluded.}  Empirical instability at
8B scale; root cause not deeply investigated.
\item \textbf{CUDA-only; NVIDIA $\geq 8$~GB VRAM required.}
\end{itemize}

% ───────────────────── §10 Future ─────────────────────
\section{Future Work}
\label{sec:future}
The highest-priority follow-up is a cross-hardware sweep on RTX~4090, A100,
and H100 to validate the predictions in \Cref{tab:predict}, paired with
Nsight Compute counter traces to confirm or refute the L2 mechanism.
Per-matrix bases ($\Pmat_Q,\Pmat_K,\Pmat_V$) would close the
3--10$\times$ Eckart--Young gap at the cost of $3\times$ projection storage.
FFN compression requires structure-aware methods (block-diagonal
factorisation, input-adaptive sparsity \cite{elhage2022superposition},
or CPU-offload) since global SVD is unacceptably lossy.  Quality recovery via
LoRA-style \cite{hu2021lora} fine-tuning on top of the projected weights
should bring the +13.30\% PPL gap toward zero.

% ───────────────────── §11 Reproducibility ─────────────────────
\section{Reproducibility}
A complete reproduction package is at
\url{https://github.com/NagusameCS/HyperTensor/blob/main/repro/REPRODUCE.md}
with expected output CSVs in
\texttt{repro/expected\_outputs/}.  Throughput tolerance is $\pm 5\%$
(GPU clock variance); PPL is deterministic to four decimal places.

\begin{lstlisting}[language=bash,caption={Headline reproduction.}]
# baseline
geodessical <model.gguf> -n 256 --temp 0 -p "Write a sorting algorithm in Python"
# GRC k=1536
geodessical <model.gguf> -n 256 --temp 0 -p "..." \
  --axex-compress --axex-attn-only --axex-skip-o \
  --axex-weight-pca --axex-compress-rank 1536
# full benchmark harness (~60 min)
./scripts/benchmark_whitepaper_finalize.ps1 -CooldownSec 30
\end{lstlisting}

\section*{Acknowledgements}
The author thanks the open-source \texttt{llama.cpp} \cite{ggerganov2023gguf}
and GGUF k-quants \cite{ggerganov2023kquants} communities, whose work
provided the Q4\_K\_M baseline this study sits on top of.

\appendix

\section{Extended empirical detail}
\label{app:extended}

This appendix collects the full measurement detail behind
\Cref{sec:results,sec:cachefit,sec:stats} for reviewers who want to
audit the headline numbers. Validated reference pack:
\texttt{benchmarks/whitepaper\_pack\_20260427\_121815/}.

\subsection{Hardware, model, and runtime configuration}

\begin{table}[h]\centering\footnotesize
\caption{Test hardware. Single configuration; cross-hardware transfer is
explicitly listed as a limitation.}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Component} & \textbf{Specification}\\
\midrule
GPU                 & NVIDIA GeForce RTX~4070 Laptop (Ada AD106)\\
GPU driver          & 595.79\\
GPU VRAM            & 8\,188\,MiB GDDR6 (256\,GB/s, 128-bit $\times$ 16\,Gbps)\\
GPU L2 cache        & \textbf{32\,MB}\\
GPU peak FP32       & 40\,TFLOPS (theoretical)\\
GPU sustained TDP   & ${\sim}109$\,W during decode (no laptop power.limit)\\
PCIe link           & Gen~4 $\times$8\\
CPU                 & AMD Ryzen~9 7940HS, 8c/16t, 4.0\,GHz base\\
System RAM          & 32\,GB DDR5-5200\\
Storage             & 2$\times$Kingston SNV2S 2\,TB NVMe\\
OS                  & Windows host, CUDA backend\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]\centering\footnotesize
\caption{Model and runtime under test.}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Property} & \textbf{Value}\\
\midrule
Model               & \texttt{bartowski/Meta-Llama-3.1-8B-Instruct-GGUF}\\
Quantisation        & Q4\_K\_M (GGUF v3)\\
File size           & 4.583\,GB ($4{,}920{,}739{,}232$\,bytes)\\
Architecture        & LLaMA, 32 layers, $d{=}4096$, 32 heads, 8 KV groups, $d_h{=}128$\\
Vocab               & 128\,256 BPE tokens\\
Parameters          & 8\,310\,M\\
Default context     & 8\,192 tokens\\
\midrule
Runtime             & Geodessical v0.6.0 ``Synapse''\\
Binary              & \texttt{build\_host/geodessical.exe}, 1.14\,MB\\
OpenBLAS DLL        & 48.63\,MB\\
Total artefacts     & 69.2\,MB\\
Backend             & CUDA 12.4 (Zig-CC build)\\
Worker threads      & 15 + 1 BSP (16 total)\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Storage footprint}

\begin{table}[h]\centering\footnotesize
\caption{On-disk artefacts. Minimum footprint for inference (model + one
$\Wmat_\text{proj}$ cache + runtime) is ${\sim}5.75$\,GB. The
$\Wmat_\text{proj}$ cache is computed once per (model, $k$) pair and reused.}
\begin{tabular}{@{}lrl@{}}
\toprule
\textbf{Artefact} & \textbf{Size} & \textbf{Notes}\\
\midrule
Model file (Q4\_K\_M GGUF)                  & 4.583\,GB   & HuggingFace blob, read-only\\
$\Wmat_\text{proj}$ cache, $k{=}1536$       & 1\,092.7\,MB & $32{\times}3$ slots, $(\Pmat_k,\widetilde\Wmat)$\\
$\Wmat_\text{proj}$ cache, $k{=}1024$       & 1\,092.7\,MB & Same shape, different eigenbasis\\
$\Wmat_\text{proj}$ cache, $k{=}2048$ req.\ & 1\,456.8\,MB & Cap clamps to $k{=}1536$ at runtime\\
KV-cache snapshot (\texttt{ott\_hs\_cache}) & 1\,024.0\,MB & one-decode hidden state\\
Runtime + OpenBLAS                          & 69.2\,MB     & complete inference stack\\
\bottomrule
\end{tabular}
\end{table}

First-run calibration time (CPU-bound truncated SVD over 32 layers $\times$
3 weight sets, dequantised Q4\_K\_M\,$\to$\,fp32) is 60--120\,s on the
Ryzen~9 7940HS; warm cache hit elides this entirely.

\subsection{VRAM profile}

\begin{table}[h]\centering\footnotesize
\caption{VRAM during decode, measured via \texttt{nvidia-smi dmon} at 1\,Hz.
Idle baseline (OS, display, background) is ${\sim}1\,136$\,MiB. Both
configurations sit comfortably inside the 8\,188\,MiB budget.}
\begin{tabular}{@{}lrrr@{}}
\toprule
\textbf{Stage} & \textbf{Baseline (MiB)} & \textbf{GRC $k{=}1536$ (MiB)} & $\Delta$ \\
\midrule
OS / display idle      & ${\sim}1\,136$        & ${\sim}1\,136$        & ---\\
Post-model upload      & ${\sim}5\,812$        & ${\sim}5\,812$        & ---\\
Active decode (sust.)  & 6\,695                & 6\,702--6\,731        & $+7$ to $+36$\\
Peak observed          & 6\,695                & 6\,731                & $+36$\\
Headroom               & ${\sim}1\,493$        & ${\sim}1\,457$        & ---\\
\bottomrule
\end{tabular}
\end{table}

Runtime-reported model load: 226 weight tensors (4\,684\,MB) +
scratch arena (513\,MB) + KV cache and activations at 8K context
(${\sim}615$\,MB). The 1.09\,GB on-disk $\Wmat_\text{proj}$ is only
\emph{partially} resident in VRAM during decode --- the GPU-resident working
set uses only the active layer's projected matrices, hence the small +36\,MiB
delta despite the large on-disk artefact.

\subsection{Power draw}

\begin{table}[h]\centering\footnotesize
\caption{GPU power profile. Both modes reach the same 103--109\,W sustained
decode envelope; GRC and baseline are bandwidth-saturated alike. PCA
calibration is CPU-bound and idles the GPU at 13--14\,W on first run only.}
\begin{tabular}{@{}lrrrr@{}}
\toprule
\textbf{Phase} & \textbf{GPU power (W)} & \textbf{GPU clock (MHz)} & \textbf{Util.} & \textbf{Temp.\ (°C)}\\
\midrule
Idle                       & 1.9--2.3   & 210         & 4--5\%     & 38--39\\
Model load (VRAM upload)   & 15.8--15.9 & 1\,980      & 35--36\%   & 39--41\\
PCA calib (CPU-bound, 1st run)
                           & 13--14     & 1\,980      & 0--1\%     & 41\\
Decode (sustained)         & 103--109   & 2\,235      & 97--100\%  & 59--61\\
Cooldown                   & 2--16      & 210--2\,235 & 0--3\%     & 39--42\\
\bottomrule
\end{tabular}
\end{table}

Per-token efficiency (baseline TpF report): HBM bandwidth
174--179\,GB/s (68--70\% of 256\,GB/s peak), compute 588--606\,GFLOPS
(1.47--1.51\% of 40\,TFLOPS peak), $\sim 4.9$\,GB loaded per token.
The $<\!2\%$ compute / $\sim\!70\%$ bandwidth split confirms the
Roofline reading: this is a memory-bound regime where bytes-saved is the
right axis to optimise.

\subsection{Full spectral analysis (rank required for 95\% energy)}

We computed the full SVD of every attention and FFN weight in five layers
($L\in\{0,7,15,23,31\}$) of Llama-3.1-8B-Instruct after Q4\_K\_M
dequantisation. The rank required to capture $95\%$ of $\norm{\Wmat}_F^2$:

\begin{table}[h]\centering\footnotesize
\caption{Per-matrix $k_{95}$ across five layers. The compression policy
``attention only'' is justified by construction --- FFN matrices have
near-uniform spectra and would need ${\sim}3.4\times$ the rank for the
same energy capture.}
\begin{tabular}{@{}lrrl@{}}
\toprule
\textbf{Matrix} & \textbf{Range $k_{95}$} & \textbf{Mean $k/d$} & \textbf{Status at GRC $k{=}1024$}\\
\midrule
$\Wmat_Q$    & 635--1947  & 0.32 & within target\\
$\Wmat_K$    & 253--783   & 0.13 & well within target\\
$\Wmat_V$    & 663--800   & 0.18 & within target\\
$\Wmat_O$    & 1536--2104 & 0.45 & marginal (excluded for stability)\\
FFN gate     & 3263--3475 & 0.83 & far exceeds GRC rank\\
FFN up       & 3398--3510 & 0.85 & far exceeds GRC rank\\
FFN down     & 3407--3525 & 0.85 & far exceeds GRC rank\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Confidence-interval pack (12-rep)}

\begin{table}[h]\centering\footnotesize
\caption{Per-class CI bounds. GRC variance is ${\sim}6\times$ baseline ---
$\pm 2$\,tok/s vs.\ $\pm 0.3$\,tok/s --- because L2 residency varies with
prompt-induced access patterns. Worst-case lower-95\% bound is 86\% of
baseline, well above the 67\% gate.}
\begin{tabular}{@{}lrrrr@{}}
\toprule
\textbf{Prompt class} & \textbf{Baseline (tok/s)} & \textbf{GRC (tok/s)} & \textbf{Mean retention} & \textbf{Lower-95\%}\\
\midrule
coding/256     & $35.68\!\pm\!0.35$ & $34.86\!\pm\!2.02$ & 97.70\% & \textbf{86.60\%}\\
reasoning/256  & $35.58\!\pm\!0.31$ & $35.22\!\pm\!2.42$ & 98.99\% & \textbf{85.64\%}\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Validation-gate summary}

The whitepaper-pack harness emits seven automated gates evaluated by
\texttt{scripts/validation\_cycle.ps1}; all seven pass on the reference pack.

\begin{table}[h]\centering\footnotesize
\caption{Validation gates on pack \texttt{whitepaper\_pack\_20260427\_121815}.}
\begin{tabular}{@{}lrrl@{}}
\toprule
\textbf{Gate} & \textbf{Threshold} & \textbf{Measured} & \textbf{Result}\\
\midrule
$k{=}1024$ decode retention             & $\geq 95\%$ & 106.27\% & PASS\\
$k{=}1536$ decode retention             & $\geq 75\%$ &  97.55\% & PASS\\
$k{=}1536^\dagger$ warmup-proxy decode  & $\geq 75\%$ & 101.04\% & PASS\\
$k{=}1536^\dagger$ warmup-proxy prefill & $\leq 225\%$& 108.48\% & PASS\\
coding/256 lower-95 retention           & $\geq 67\%$ &  86.60\% & PASS\\
reasoning/256 lower-95 retention        & $\geq 67\%$ &  85.64\% & PASS\\
PPL delta (WikiText-2)                  & $\leq +15\%$&  +13.30\% & PASS\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Benefits and limitations catalogue}

\begin{table}[h]\centering\footnotesize
\caption{Honest accounting of GRC's benefits and limits, distilled from the
v0.6.0 technical report. Severities reflect deployment impact.}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Dimension} & \textbf{Sign} & \textbf{Severity}\\
\midrule
No calibration data needed (pure weight geometry)  & $+$ & Major\\
Super-baseline at $k{=}1024$ (106.27\% decode)     & $+$ & Major\\
Near-lossless throughput at $k{=}1536$             & $+$ & Moderate\\
Deterministic, cacheable, sign-canonical basis     & $+$ & Moderate\\
Sub-second warm startup (small models)             & $+$ & Moderate\\
Zero training infrastructure                       & $+$ & Major\\
Modular (Q/K/V only; FFN untouched)                & $+$ & Moderate\\
\midrule
$+13.30\%$ WikiText-2 PPL at $k{=}1536$            & $-$ & Major\\
Prefill 8--15\% slower (sequential path)           & $-$ & Moderate\\
\texttt{AXEX\_MANIFOLD\_K\_MAX} hard cap at 1536   & $-$ & Moderate\\
Minimal VRAM reduction ($+36$\,MiB delta)          & $-$ & Moderate\\
First-run cold start 60--120\,s                    & $-$ & Moderate\\
GRC throughput variance ${\sim}6\times$ baseline   & $-$ & Moderate\\
No power savings (same 109\,W TDP)                 & $-$ & Mild\\
$\Wmat_O$ excluded; FFN excluded                   & $-$ & Mild\\
Single hardware, single model validated            & $-$ & Major\\
CUDA-only (no ROCm/Metal/CPU fallback)             & $-$ & Mild\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Methodological gaps}

\begin{table}[h]\centering\footnotesize
\caption{Acknowledged methodological gaps. Items 1, 3, 4 require
infrastructure outside the scope of this independent project; item~2 is on
the near-term roadmap.}
\begin{tabular}{@{}p{0.28\linewidth}p{0.30\linewidth}p{0.32\linewidth}@{}}
\toprule
\textbf{Gap} & \textbf{Missing} & \textbf{Why it matters}\\
\midrule
Direct L2 cache-hit measurement & No Nsight Compute counter trace; cache-fit argued from access-pattern arithmetic. & Alternative micro-arch explanations (register pressure, scheduler) cannot be ruled out without counters.\\
Task-level evaluations & Only WikiText-2 PPL; no MMLU/GSM8K/HumanEval/IFEval. & $+13.30\%$ PPL is a structural signal, not a behavioural one.\\
Head-to-head vs.\ AWQ / GPTQ / SmoothQuant & No direct A/B on identical hardware; we sit on the same Q4\_K\_M baseline. & Production-scale ranking unverifiable without compatible-runtime baselines.\\
Cross-hardware validation & All measurements on RTX~4070 Laptop; \Cref{tab:predict} is calculated, not measured. & Cache-fit cannot be claimed as general --- only as observed on this GPU.\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Phase status}

\begin{table}[h]\centering\footnotesize
\caption{Project phase status as of v0.6.0.}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Phase} & \textbf{Objective} & \textbf{Status}\\
\midrule
1 & Eliminate measurement instability       & \textbf{Complete} (root cause: thermal throttling; 30\,s cooldown fix)\\
2 & Validate under locked protocol          & \textbf{Complete} (7/7 gates pass)\\
3 & Cross-model transfer (Mistral-7B-v0.1)  & In progress\\
3 & Cross-hardware transfer                 & Blocked (single-GPU constraint)\\
4 & External reproduction package           & \textbf{Complete} (\texttt{repro/REPRODUCE.md})\\
\bottomrule
\end{tabular}
\end{table}

\subsection{Practical implication}

A deployer can select compression rank as a latency/quality dial:
\begin{itemize}
\item $k{=}1024$: maximum throughput ($+6\%$), structural PPL collapse on
this model (GQA $\Wmat_K,\Wmat_V$ already at rank $\leq 1024$).
\item $k{=}1536$: near-lossless throughput ($-2.45\%$), measurable but
deployment-acceptable quality cost ($+13.30\%$ PPL).
\item $k{=}2048$ requested (silently capped to 1536 by
\texttt{AXEX\_MANIFOLD\_K\_MAX}): warmup proxy at parity with $k{=}1536$.
\end{itemize}
The headline reading is that attention compression with GRC is no longer a
strict trade-off; it is a tunable parameter where one operating point
exceeds baseline and another is statistically indistinguishable from it.

\printbibliography

\end{document}
