\documentclass[11pt]{article}
\usepackage{hypertensor}
\graphicspath{{figures/}}
\addbibresource{refs.bib}

\title{Light Distillation for Calibration-Permitted Low-Rank \\
       Attention Compression\\
       {\large An optional companion to GRC}}
\author{HyperTensor Research}
\date{April 2026}

\begin{document}
\maketitle
\hypertensorbanner

\begin{abstract}
\noindent
Paper~A introduced GRC, a calibration-free shared-basis low-rank
projection of attention weights that is fast on consumer GPUs
($+6.27\%$ throughput at $k{=}1024$ on Llama-3.1-8B Q4\_K\_M, RTX~4070
Laptop) but pays a measured $+13.30\%$ perplexity penalty at the
quality-preserving rank $k{=}1536$. This companion paper introduces an
optional, opt-in distillation step that recovers a configurable
fraction of that penalty in exchange for a few hundred forward passes
on a small calibration corpus. The contribution is a protocol, not new
math: per-layer rank-$r$ LoRA adapters \cite{hu2021lora} fit on the
teacher-student logit residual after GRC projection, with sink-channel
exemption \cite{sun2024massive,xiao2024streamingllm} as an orthogonal
quality lever. We describe the design, the runtime path that allows
distillation to remain optional (the calibration-free GRC path stays
the default and the distilled adapters are mergeable), and the
empirical protocol that an EC2 GPU run should follow. \emph{Status:
this paper is a scaffold. Empirical results, baselines and ablations
will be filled in once the distillation runner has executed on
appropriate hardware. Section~\ref{sec:status} tracks the gating
items.}
\end{abstract}

\clearpage
\tableofcontents
\clearpage

\section{Motivation}
\label{sec:motivation}

Paper~A is deliberately calibration-free: GRC requires no data, no
fine-tuning, no per-deployment configuration beyond a single rank knob.
That property is valuable for the deployment surface it targets
(consumer GPUs, ad-hoc compression, no MLOps stack), and the empirical
trade is that there is a quality floor set by the joint Gram spectrum
itself. On Llama-3.1-8B-Instruct the floor at $k{=}1024$ is roughly the
bottom $9\%$ of joint Q+K+V energy ($k_{95}{\approx}1682$), and the
realised perplexity penalty is $+13.30\%$ at $k{=}1536$ on WikiText-2
\cite{merity2017wikitext}. Paper~A's Limitations section decomposes
that penalty into intrinsic floor and reducible implementation choice.

This paper exercises the largest reducible lever, light distillation,
behind an opt-in flag. The design constraint is that turning it on
must not break Paper~A's runtime path; turning it off must reproduce
Paper~A's numbers exactly.

\section{Method}
\label{sec:method}

\subsection{Phase 1: GRC projection (unchanged from Paper~A)}
For each attention block, build the shared basis $\Pmat_t$ from the
joint Gram of $\Wmat_Q, \Wmat_K, \Wmat_V$ and form the projected
weights $\Wmat_X^\prime = \Wmat_X \Pmat_k \Pmat_k^\top$. Optionally,
exempt the top-$T$ highest-L2 columns from projection (sink-aware GRC,
see Paper~A Table~6 / \texttt{sink\_channel\_pilot.json}). This phase
is pure NumPy and runs on any host.

\subsection{Phase 2: Teacher-student LoRA correction (new)}
Fit per-layer rank-$r$ LoRA adapters
$\Amat_X \in \mathbb{R}^{m \times r}$, $\Bmat_X \in \mathbb{R}^{r \times d}$
for $X \in \{Q, K, V\}$ such that
$$
\hat{\Wmat}_X = \Wmat_X^\prime + \Amat_X \Bmat_X
$$
minimises a teacher-student logit MSE over a small calibration corpus
(target: 5{,}000 to 10{,}000 tokens, sequence length 512, batch 8;
500-step schedule on AdamW with $\eta{=}10^{-4}$, weight decay $0$,
cosine LR with warmup-100). The teacher is the original uncompressed
model; the student is the projected model with frozen
$\Wmat_X^\prime$ and trainable $(\Amat_X, \Bmat_X)$. Only the
attention QKV adapters are fit; output projection $\Wmat_O$ and FFN
weights remain at baseline. This holds the parameter budget below
$3rdL = 3 \cdot 8 \cdot 4096 \cdot 32 \approx 3.1\,\text{M}$ trainable
weights for $L{=}32$ layers and $r{=}8$, two orders of magnitude below
full LoRA fine-tuning.

\subsection{Phase 3: Merge and ship}
The trained LoRA factors are folded into the projected weights
($\hat{\Wmat}_X = \Wmat_X^\prime + \Amat_X \Bmat_X$) and re-quantised
to Q4\_K\_M via the standard llama.cpp quantize path. The runtime
loads the resulting GGUF unchanged; no runtime code changes are
needed. If the user disables distillation, Paper~A's pipeline runs
unmodified.

\section{Implementation status}
\label{sec:status}

\begin{itemize}
\item Done. Phase~1 reference implementation in
\texttt{scripts/analysis/sink\_channel\_grc.py} and
\texttt{scripts/grc\_distill.py} (NumPy; CPU-only).
\item Done. Sink-channel pilot (Paper~A Table~6) showing
calibration-free sink exemption alone yields 1 to 3\% relative
reconstruction-error improvement at $k\in\{512,1024,1536\}$.
\item Pending. Phase~2 PyTorch runner. Requires a GPU host
with at least one A100 / L40S / H100 to hold the teacher in BF16 with
gradient activations. EC2 \texttt{g5.xlarge} or \texttt{g6.xlarge} is
sufficient for an 8B model.
\item Pending. WikiText-2 PPL evaluation on the merged
artifact. The expected operating regimes:
\begin{enumerate}
  \item Vanilla GRC at $k{=}1024$: $+13.30\%$ PPL (Paper~A baseline).
  \item Sink-aware GRC at $k{=}1024$, $T{=}32$: predicted $+12.5\%$
        to $+13.0\%$ (small effect; the calibration-free magnitude proxy
        weakly correlates with runtime sink channels).
  \item Distilled GRC at $k{=}1024$, $r{=}8$, 5{,}000-token corpus:
        target $+5\%$ to $+8\%$ PPL (50 to 60\% gap closure).
  \item Distilled sink-aware GRC at $k{=}1024$, $T{=}32$, $r{=}8$:
        target $+3\%$ to $+5\%$.
\end{enumerate}
These targets are predictions, not measurements. The paper will be
updated with measured numbers once the runner has executed.
\item Pending. Throughput re-measurement on the distilled
artifact. Hypothesis: throughput is unchanged from vanilla GRC,
because the merged adapters do not alter the kernel-fusion path
(\Cref{sec:fusion-fit}). If measurement contradicts the hypothesis,
that becomes a result.
\end{itemize}

\section{Why distillation does not break the fusion path}
\label{sec:fusion-fit}

Paper~A established that the GRC throughput improvement comes from
kernel fusion: three $\texttt{gemv\_q4\_k}$ calls collapse into a
fused $\texttt{gemv\_dual\_q8\_0}$ trio operating on
$k$-dimensional intermediates. The fusion-viability condition is that
the projected weights $\Wmat_X^\prime$ have rank $k$. The merged
weights $\hat{\Wmat}_X = \Wmat_X^\prime + \Amat_X \Bmat_X$ in general
have rank $k + r$ (when $r \le m - k$), but the $\Amat_X \Bmat_X$
term has rank exactly $r$ and is small; the runtime can either
(a) treat $\hat{\Wmat}_X$ as a full-rank matrix and lose fusion (the
$+6.27\%$ throughput gain reverts), (b) keep $\Wmat_X^\prime$ on the
fused path and run the LoRA correction as a separate small
matmul-add (cost $\sim r/k$ of the main GEMV; negligible at $r{=}8$,
$k{=}1024$), or (c) re-quantise the merged $\hat{\Wmat}$ to Q4\_K\_M
at rank $k$ via a second SVD pass and accept the merge-loss. Default
recommendation is (b); (c) is reserved for the lowest-VRAM target
where adapter storage matters.

\section{Reproducibility}
\label{sec:repro}

The reference Phase-1 code lives in
\texttt{scripts/analysis/sink\_channel\_grc.py} and
\texttt{scripts/grc\_distill.py}. The Phase-2 runner will live in
\texttt{scripts/grc\_distill\_runner.py} (PyTorch + transformers) and
the EC2 driver in \texttt{scripts/ec2\_paperE\_distill/}. A complete
empirical reproduction will be published alongside the populated
results once the runner has executed.

\printbibliography

\end{document}
