% HyperTensor (Part XXV): Gauge Orbit Manifold
% Blueprint — Yang-Mills mass gap via spectral geometry on gauge orbits.
\documentclass{article}
\input{hypertensor.sty}

\title{HyperTensor (Part~XXV): The Gauge Orbit Manifold\\
and the Yang--Mills Mass Gap}
\author{NagusameCS}
\date{May 2026}

\begin{document}
\maketitle

\begin{abstract}
The Yang--Mills existence and mass gap problem asks: for any compact simple
gauge group $G$, does the quantum Yang--Mills theory exist on $\mathbb{R}^4$
and exhibit a positive mass gap $\Delta > 0$?  Paper~XXV reframes this as a
spectral geometry problem on the \textbf{Gauge Orbit Manifold (GOM)}: a
$k$-manifold whose points are gauge equivalence classes of connections and
whose Laplacian spectrum encodes the mass spectrum of the quantum theory.

The key insight: the mass gap $\Delta$ is the \textbf{first nonzero eigenvalue}
of the Laplace--Beltrami operator $\Delta_{\mathcal{M}}$ on the GOM.  Proving
$\Delta > 0$ reduces to proving that the GOM is compact with a spectral gap.
HyperTensor's $k$-manifold framework provides direct computational access to
the Laplacian on learned manifolds, offering a new attack vector on this
50-year-old problem.
\end{abstract}

\section{The Millennium Problem}
\label{sec:ym-problem}

The Yang--Mills equations generalize Maxwell's equations to non-abelian gauge
groups.  The classical action is:
\begin{equation}
S_{\mathrm{YM}}[A] = \frac{1}{4g^2} \int \mathrm{Tr}(F_{\mu\nu}F^{\mu\nu}) \, d^4x
\end{equation}
where $F_{\mu\nu} = \partial_\mu A_\nu - \partial_\nu A_\mu + [A_\mu, A_\nu]$
is the field strength tensor for the gauge connection $A_\mu$ taking values
in the Lie algebra $\mathfrak{g}$.

The millennium problem asks for a rigorous construction of the quantum theory
on $\mathbb{R}^4$ with a \textbf{mass gap}: the existence of a positive lower
bound $\Delta > 0$ on the energy of all excited states above the vacuum.

\section{The Gauge Orbit Manifold}
\label{sec:gom}

\subsection{Gauge Orbits as Manifold Points}

A gauge connection $A_\mu(x)$ is defined only up to gauge transformations:
\begin{equation}
A_\mu \mapsto g A_\mu g^{-1} + g \partial_\mu g^{-1}, \quad g(x) \in G
\end{equation}

The \textbf{gauge orbit} $[A]$ is the equivalence class of all connections
related by gauge transformations.  The space of gauge orbits is the true
configuration space of the quantum theory.

The GOM $\mathcal{M}_{\mathrm{YM}}$ embeds gauge orbits into a $k$-manifold
via a learned encoder $\Phi: \mathcal{A}/\mathcal{G} \to \mathcal{M}_{\mathrm{YM}}$.

\subsection{The Mass Gap as Spectral Gap}

On the GOM, the Hamiltonian of the quantum Yang--Mills theory corresponds to
the \textbf{Laplace--Beltrami operator} $\Delta_{\mathcal{M}}$:
\begin{equation}
\hat{H} \longleftrightarrow -\frac{\hbar^2}{2} \Delta_{\mathcal{M}} + V_{\mathrm{eff}}
\end{equation}

The vacuum state $|0\rangle$ is the ground eigenfunction of $\Delta_{\mathcal{M}}$
(constant function, eigenvalue $0$).  The mass gap $\Delta$ is the \textbf{first
nonzero eigenvalue}:
\begin{equation}
\Delta = \lambda_1(\Delta_{\mathcal{M}}) > 0
\end{equation}

Proving the Yang--Mills mass gap is equivalent to proving that the GOM has a
\textbf{spectral gap}---a positive lower bound on $\lambda_1$.

\section{Spectral Geometry on the GOM}
\label{sec:spectral}

\subsection{The Weyl Law on Learned Manifolds}

The asymptotic distribution of Laplacian eigenvalues on a compact $k$-manifold
follows Weyl's law:
\begin{equation}
N(\lambda) = \frac{\omega_k \mathrm{vol}(\mathcal{M})}{(2\pi)^k} \lambda^{k/2} + O(\lambda^{(k-1)/2})
\end{equation}

The first nonzero eigenvalue $\lambda_1$ is bounded below by the
\textbf{Cheeger constant} $h(\mathcal{M})$:
\begin{equation}
\lambda_1 \geq \frac{h(\mathcal{M})^2}{4}
\end{equation}

where $h(\mathcal{M}) = \inf \frac{\mathrm{area}(\partial S)}{\mathrm{vol}(S)}$
over all submanifolds $S \subset \mathcal{M}$.  The Cheeger constant is a
geometric isoperimetric quantity that can be estimated from the learned metric.

\subsection{Computing $\lambda_1$ on a Learned Manifold}

Given the GOM embedding $\Phi$, we can:
\begin{enumerate}
  \item Sample points $\mathbf{x}_i = \Phi([A_i])$ from the manifold.
  \item Construct the graph Laplacian $L_{ij}$ from pairwise geodesic distances.
  \item Compute eigenvalues of $L$; the first nonzero eigenvalue estimates
    $\lambda_1$.
  \item Prove a lower bound on $\lambda_1$ using Cheeger's inequality and
    the learned metric tensor.
\end{enumerate}

\section{The Instanton Gas and the Mass Gap}
\label{sec:instantons}

\subsection{Instantons as Geodesic Loops}

Yang--Mills instantons are finite-action solutions to the Euclidean field
equations.  On the GOM, instantons correspond to \textbf{closed geodesics}
connecting different vacua (pure gauge configurations at infinity).

The instanton gas picture (Callan--Dashen--Gross) suggests that the mass gap
arises from instanton--anti-instanton interactions.  On the GOM, this
translates to the statement that \textbf{the fundamental group
$\pi_1(\mathcal{M}_{\mathrm{YM}})$ is nontrivial}, and the mass gap is the
energy required to traverse a non-contractible loop.

\subsection{Curvature and Confinement}

The Wilson loop expectation value $\langle W(C) \rangle$ measures the
holonomy around a closed curve $C$.  Area-law behavior ($\langle W \rangle
\sim e^{-\sigma \cdot \mathrm{area}}$) signals confinement.  On the GOM,
area-law corresponds to \textbf{positive sectional curvature} in the
directions corresponding to spatial Wilson loops.

\section{TEH and the Mass Gap}
\label{sec:teh-ym}

\subsection{The Vacuum as a Fixed Point}

The vacuum state corresponds to the ``flat connection'' gauge orbit
$[A_\mu = 0]$.  On the GOM, we train the manifold such that the vacuum is
a \textbf{stable fixed point} of the gradient flow.  TEH are engineered
around the vacuum boundary---the region where fluctuations become large
enough to create massive excitations.

The mass gap then has a geometric interpretation: \textbf{the TEH radius
around the vacuum point.}  Excitations within the TEH radius are vacuum
fluctuations (massless); excitations that cross the TEH acquire mass
$\Delta$.

\section{Proof Strategy}
\label{sec:proof-strategy}

\begin{enumerate}
  \item \textbf{Train GOM on lattice gauge configurations.}  Use
    Monte Carlo data from lattice QCD ($\beta = 6/g^2$) to embed gauge
    configurations on the GOM.
  \item \textbf{Verify compactness.}  Prove that the learned GOM metric
    is complete and the manifold is compact (finite volume).  This
    follows if the gauge group $G$ is compact and the lattice is finite.
  \item \textbf{Compute spectral gap.}  Estimate $\lambda_1$ from the
    graph Laplacian; verify that $\lambda_1 > 0$ and stable under
    refinement of the lattice.
  \item \textbf{Prove lower bound.}  Use Cheeger's inequality and the
    learned metric to prove $\lambda_1 \geq \Delta_0 > 0$ as the lattice
    spacing $a \to 0$.
  \item \textbf{Continuum limit.}  Take $a \to 0$ while holding the
    physical mass gap fixed; verify that $\lambda_1$ approaches a finite
    positive limit.
\end{enumerate}

\section{Scaling Requirements}
\label{sec:scaling}

\begin{itemize}
  \item \textbf{Lattice:} $N^4$ with $N = 16$--$64$ ($\sim 10^5$--$10^7$
    gauge links)
  \item \textbf{Manifold:} $k = 1024$--$4096$, $d = 4096$--$16384$
  \item \textbf{Training data:} $10^5$--$10^7$ lattice configurations
  \item \textbf{Model:} 7B--70B parameters
  \item \textbf{Key challenge:} The gauge symmetry must be exactly preserved
    by the embedding; otherwise the GOM conflates gauge-equivalent states.
\end{itemize}

\section{Conclusion}
\label{sec:conclusion}

Paper~XXV demonstrates that the Yang--Mills mass gap problem admits a spectral
geometry formulation on a learned gauge orbit manifold.  The mass gap is the
first nonzero Laplacian eigenvalue; proving $\Delta > 0$ is a question about
the Cheeger constant of the GOM.

With AGT (XVI--XVII) for the Riemann Hypothesis, CCM (XIX) for P vs NP, HSM
(XXII) for Navier--Stokes, and GOM (XXV) for Yang--Mills, four of the seven
millennium problems now have geometric formulations within the HyperTensor
framework.  The pattern is unmistakable: \textbf{the deepest problems in
mathematics and physics are geometric questions about learned manifolds.}

\end{document}
