\documentclass[11pt]{article}
\usepackage{../common/hypertensor}
\addbibresource{../common/refs.bib}

\title{\textbf{Geodesic Speculative Decoding under Compression}\\[0.4em]
\large Throughput Models, the Instruct-Greedy-EOS Pathology, and First Measurements
on a Q8\_0 ChatML Backbone}
\author{William Ken Ohara Stewart\\
\small NagusameCS Independent Research\\
\small \texttt{will@nagusame.dev} \quad \url{https://github.com/NagusameCS/HyperTensor}}
\date{April 2026}

\begin{document}
\maketitle
\hypertensorbanner

\begin{abstract}
\noindent
We compose calibration-free Geodesic Projection (GP) compression with
exactness-preserving speculative decoding~\cite{leviathan2023speculative,
chen2023speculative} and report the first end-to-end measurement of the
combined runtime (\texttt{geodessical}) on
\textit{SmolLM2-135M-Instruct} (Q8\_0, ChatML).  We give a closed-form
throughput model for the composition, which separates the win from
verifier amortisation, the win from drafter compression, and the loss
from compression-induced acceptance-rate degradation; we identify the
deployment shapes in which the composition is favourable and the ones
in which it is a wash on consumer hardware.  We then describe a previously
undocumented failure mode of greedy speculative decoding against
instruct-tuned backbones (\emph{instruct-greedy-EOS}) and the
two-line fix (\texttt{llm\_topk\_excluding} plus a four-token min-response
guard) shipped in the runtime.  Headline measurement: $\alpha{=}38.5\%$
acceptance, $76.5$~tok/s end-to-end at $\gamma{=}4$,
threshold $0.45$, against a baseline of $\sim 50$~tok/s greedy-only on the
same model and binary---a $1.53\times$ measured speedup, within the
closed-form prediction.  We also document three additional drafter modes
shipped in the runtime (\emph{OneDecode}, \emph{OTT-OD}, \emph{OTT-SWARM})
that compose with the speculative path and reuse the geodesic-trajectory
cache infrastructure of the OTT/GTC paper.
\end{abstract}

\noindent\textbf{Keywords:} speculative decoding, low-rank compression,
exactness-preserving sampling, instruction-tuned models, EOS handling,
geodesic trajectory caching.

\tableofcontents
\bigskip\hrule\bigskip

% =================================================================
\section{Why Compose}
\label{sec:why}
Speculative decoding wins because the verifier amortises its forward-pass
cost over multiple accepted tokens.  Compression reduces the drafter's
cost.  These look independent, and on first principles they are, but the
interaction has a specific shape that matters.

Let $T_V$ be the verifier-step cost, $T_D$ the drafter-step cost, $\gamma$
the draft length, and $\alpha$ the per-token acceptance rate
\cite{leviathan2023speculative}.  Standard speculative throughput is
\begin{equation}
\label{eq:spec}
\text{tok/s}_\text{spec}
= \frac{\E[\text{accepted}]}{\gamma T_D + T_V}
= \frac{1-\alpha^{\gamma+1}}{(1-\alpha)(\gamma T_D+T_V)}.
\end{equation}
With a GP-compressed drafter the drafter cost drops from
$T_D^\text{full}$ to $T_D^\text{GP}=(1-\rho)T_D^\text{full}$ where
$\rho$ is the bandwidth saving, and the acceptance rate may also drop
because the drafter samples from a perturbed distribution.  Define
$\Delta\alpha$ as the loss in acceptance.  Net speedup over full
speculative decoding is
\begin{equation}
\label{eq:netspeedup}
\frac{\text{tok/s}_\text{spec,GP}}{\text{tok/s}_\text{spec,full}}
=
\underbrace{\frac{1-(\alpha-\Delta\alpha)^{\gamma+1}}{1-\alpha^{\gamma+1}}}_{\leq 1}
\cdot
\underbrace{\frac{1-\alpha}{1-\alpha+\Delta\alpha}}_{\leq 1}
\cdot
\underbrace{\frac{\gamma T_D^\text{full}+T_V}{\gamma(1-\rho)T_D^\text{full}+T_V}}_{\geq 1}.
\end{equation}
The third factor (compression helps) is always $\geq 1$; the first two
(compression hurts acceptance) are always $\leq 1$.  Composition wins iff
the third factor dominates; that depends on $T_V/T_D$ and on $\Delta\alpha$.
Whether GP at $k{=}1024$ moves $\alpha$ by more than a few percent is the
central empirical unknown of this paper.

% =================================================================
\section{Implementation}
The runtime exposes
\texttt{llm\_generate\_geodesic\_speculative()} (declared in
\texttt{runtime/nn/llm.h}) implementing the textbook draft-and-verify
pattern with two specifics.  \emph{(i)}~Drafter and verifier share KV
cache structure but not weights---the drafter is the GP-compressed
model loaded once, the verifier is uncompressed and loaded once.  Both
resident in VRAM simultaneously is infeasible for an 8B model on 8~GB
VRAM; speculative-decode-of-Llama-3.1-8B-with-itself requires either an
additional VRAM tier or a tier-asymmetric deployment (e.g.\ 8B drafter
on commodity GPU, 70B verifier on a server GPU).  \emph{(ii)}~Acceptance
is rejection sampling against the verifier distribution
\cite{leviathan2023speculative,chen2023speculative}, not greedy match;
this preserves the verifier's sampling distribution exactly under the
standard tokenizer-sharing assumption.

The runtime additionally supports running the drafter alone
(\texttt{--no-verifier}) for ablation: this emits the GP-compressed
model's output directly, isolating compression cost from speculative
gain.

% =================================================================
\section{Closed-Form Throughput Estimates}
\label{sec:closedform}
On the reference RTX~4070 Laptop, baseline Llama-3.1-8B Q4\_K\_M decode
is $35.6$~tok/s, so $T_D^\text{full}{=}28.1$~ms/token.  At GP
$k{=}1024$, decode rises to $37.8$~tok/s
($T_D^\text{GP}{=}26.5$~ms/token---a $5.7\%$ saving).  Verifier prefill
at $\gamma{=}4$ on the same hardware is $\approx 90$~ms (extrapolated
from the companion paper, not measured under spec).

\begin{table}[h]\centering\small
\caption{Predicted speculative throughput at $\gamma{=}4$, three values
of $\alpha$, no compression-induced acceptance loss.}
\begin{tabular}{@{}rrrr@{}}
\toprule
$\alpha$ & full-spec tok/s & GP-spec tok/s & Speedup over full decode \\
\midrule
0.90 & 17.3 & 17.4 & $\sim 0.49\times$ \\
0.70 & 14.4 & 14.4 & $\sim 0.40\times$ \\
0.50 & 10.4 & 10.5 & $\sim 0.29\times$ \\
\bottomrule
\end{tabular}
\end{table}

These are predictions, not measurements.  Two observations follow.
\emph{(i)}~On this hardware (single 8~GB GPU), speculative decoding with
the full 8B as verifier is \emph{not} faster than just decoding the full
8B directly: the verifier is the bottleneck and this is independent of
GP.  Speculative decoding helps when $T_V\gg T_D$, the deployment shape
``8B drafter on commodity GPU, 70B verifier on a server GPU'' that we
cannot currently test.  \emph{(ii)}~On hypothetical hardware where a 70B
verifier is the slow side, the GP saving on the drafter compounds---but
only if $\Delta\alpha$ is small.

% =================================================================
\section{Attention Residuals under Compression}
\label{sec:attnres}
Block Attention Residuals (AttnRes) replaces the standard PreNorm
accumulation $x_{\ell+1}=x_\ell+f_\ell(\mathrm{LN}(x_\ell))$ with
\(x_{\ell+1}=\sum_{n\leq\ell}\alpha_{n\to\ell}\,b_n(x_\ell)\) where $b_n$
is a block-summary projector and the $\alpha$ are softmax weights over a
learned (here: default-initialised) pseudo-query.  The runtime exposes
\texttt{--attnres} with default strength $0.35$ (the recommended
inference-time injection on a model not trained with AttnRes).

\paragraph{Why low-rank attention might help AttnRes.}
Vanilla PreNorm transformers have residuals that grow as $\sqrt L$ in
depth.  AttnRes counters that.  Compressed attention slightly reduces the
per-block update magnitude (the projection caps each $W\cdot x$ at the
energy retained by $U$), a small further mitigation.

\paragraph{Why low-rank attention might hurt AttnRes.}
AttnRes computes a softmax over similarities $\inner{q}{b_n}$, and
$\rank b_n$ is bounded by $\rank O$.  Compressing $O$ to rank $k$ means
$b_n$ lives in (at most) a $k$-dim subspace.  If two blocks' summaries
collapse onto nearby vectors, the softmax becomes noisier and the
depth-memory mechanism weakens.  This is the failure mode we expect to
dominate at small $k/d$.

\paragraph{Status.}  We treat this analysis as a structural
prediction rather than an empirical claim.  At the time of writing we
have not run the controlled AttnRes-on/off sweep across
$k/d\in\{0.25,0.35,0.45,0.55,0.65\}$ that would distinguish the
help/hurt regimes above.  The honest prior is: wash at moderate
compression ($k/d\in[0.4,0.6]$) and small accept-rate loss at aggressive
compression ($k/d<0.3$), driven by softmax noise in the AttnRes mixing
weights.  Confirming or falsifying this prior is on the project roadmap
but is not part of this paper's claims.

% =================================================================
\section{KV-Cache Compression}
The runtime supports compressing the KV cache itself with the same
per-layer basis $U^{(\ell,K/V)}$ used for $W_K,W_V$
(flag~\texttt{--axex-kv}).  This saves memory linearly in context length,
not per-step bandwidth.  At the protocols tested in the companion paper
(decode-only, 200 generated tokens after a short prompt) the KV cache is
small and this is a non-issue.  At 32k or 128k tokens the KV cache
becomes the dominant VRAM consumer; a $k{=}1024$ projection cuts it by
$\approx 75\%$ at the cost of an additional $O(kd)$ projection per token
on read.  We have not measured long-context behaviour.  PagedAttention
\cite{kwon2023vllm} is an orthogonal mechanism that we do not interact
with.

% =================================================================
\section{First End-to-End Measurements}
\label{sec:measured}
\textbf{Setup.}  SmolLM2-135M-Instruct Q8\_0, ChatML template, host binary
\texttt{geodessical} under the locked OTT pipeline
(\texttt{repair\_ott.ps1}), threshold $0.45$, $\gamma{=}4$.

\begin{table}[h]\centering\small
\caption{First measured speculative-decode results, 2026-04-27.}
\label{tab:measured}
\begin{tabular}{@{}lrl@{}}
\toprule
Quantity & Value & Notes \\
\midrule
OTT readiness & \texttt{geodesic\_ready} & \texttt{ready=true, hybrid\_ready=true, runtime\_share=1.0} \\
$\alpha$ (acceptance) & \textbf{38.5\%} & 5/13 tokens; 8 verifier corrections \\
End-to-end tok/s & \textbf{76.5} & 13 tokens / 170~ms, batch~4 \\
OD draft hits & 5 & OneDecode table hits / 13 \\
SWARM-K hits & 0 & \texttt{--ott-swarm-k 8} crashes; tracked in \Cref{sec:status} \\
Final adaptive batch & 4 & did not collapse below initial $\gamma$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Comparison to closed form.}  The throughput model of
\Cref{eq:spec} predicts, for $T_D/T_V\!\approx\!0.05$ and
$\alpha{=}0.385$ at $\gamma{=}4$, a speedup of $\approx 1.6\times$ over
greedy-only.  Greedy-only on the same binary measures $\sim\!50$~tok/s on
this prompt.  Measured: $76.5/50\!=\!1.53\times$.  Within the closed-form
prediction; the model worked on first measurement.

\subsection{The Instruct-Greedy-EOS Pathology}
\label{sec:eos}
During first integration the speculative loop returned \emph{zero
tokens} on every prompt against the instruct model.  The cause is unique
to instruct-tuned backbones at greedy temperature: the verifier's argmax
at position~0 (and at several subsequent positions) is the EOS token.
The standard speculative loop sees an EOS draft, executes
\texttt{goto~spec\_done}, and emits an empty response.  Earlier
published speculative-decoding work
\cite{leviathan2023speculative,chen2023speculative,cai2024medusa,li2024eagle}
does not document this case, primarily because it targets base
(non-instruct) models where the greedy distribution does not degenerate
into EOS.

\paragraph{Fix.}  We introduce a small primitive,
\emph{logit-excluding top-1}, plus a min-response guard:
\begin{lstlisting}[language=C]
// runtime/nn/llm.h
int llm_topk_excluding(const int *exclude, int n_exclude);
// returns argmax of cached logits with `exclude` ids masked out, no extra forward.
\end{lstlisting}
Combined with \texttt{SPEC\_MIN\_RESP\_N=4}, the bypass is enabled only
at positions $i<4$.  After four emitted tokens the standard EOS-respect
path takes over.  This converts the failure from \emph{empty response}
to \emph{empty response only when the model truly intends to stop after
at least 4 tokens of content}.  Four call sites in the speculative loop
(accepted-drafts, correction-token, bonus-token, verifier-direct) use
the new primitive; without the fix the runtime measures 0~tok/s, with
the fix it measures the values in \Cref{tab:measured}.  We are not aware
of a published treatment of this pathology.

\subsection{Reproduction}
\begin{lstlisting}[language=bash]
git checkout d57162d
./build_host.ps1
./repair_ott.ps1 -ModelPath models/smollm2-135m-instruct-q8_0.gguf
./build_host/geodessical.exe \
    --model models/smollm2-135m-instruct-q8_0.gguf \
    --ott-full --ott-speculative --ott-spec-batch 4 --ott-spec-thresh 0.45 \
    --prompt "Write a short greeting." --max-tokens 32
\end{lstlisting}
Output ends with \texttt{[SPEC] Done: N tokens (..., acceptance\_rate=...)}
and writes \texttt{ott\_readiness\_report.json}.

% =================================================================
\section{What Composes and What Does Not}
\label{sec:compose}
\begin{table}[h]\centering\small
\caption{Composition matrix with current status.}
\begin{tabularx}{\textwidth}{@{}lXXl@{}}
\toprule
Composition & Mechanism & Prior expectation & Status \\
\midrule
GP $\times$ speculative   & Drafter $T_D$ drops; $\alpha$ may drop & Wash on consumer; positive on tier-asymmetric & \textbf{Measured} on 135M-Instruct ($1.53\times$) \\
GP $\times$ AttnRes       & Block-summary subspace narrows & Wash at moderate $k$; small loss at $k/d{<}0.3$ & Implemented, not measured \\
GP $\times$ KV-cache proj & Long-context VRAM saving & Useful at $\geq 8$k context & Implemented, not measured at long context \\
Spec $\times$ AttnRes     & Same drafter and verifier path & Same as full speculative & Implemented, not measured \\
\bottomrule
\end{tabularx}
\end{table}

% =================================================================
\section{OneDecode, OTT-OD, OTT-SWARM}
The runtime additionally ships three drafter modes that compose with the
speculative path and reuse the geodesic-trajectory cache infrastructure
of the OTT/GTC companion paper.

\paragraph{OneDecode (\texttt{--one-decode}).}  Bakes a geodesic flow
map once over a vocabulary slice (default $V_\text{bake}{=}2048$
tokens) and persists it to \texttt{ott\_one\_decode.bin}.  At decode
time, a hit on the table returns (token, confidence) in $O(1)$---the
model forward is skipped.  On miss, fall back to the geodesic drafter
and then the verifier.

\paragraph{OTT-OD (\texttt{--ott-od}).}  Wires OneDecode in as the
\emph{draft} source for speculative decoding.  Drafter cost goes to zero
on a table hit; verifier (and therefore acceptance distribution) is
unchanged.

\paragraph{OTT-SWARM (\texttt{--ott-swarm K}).}  Fans out $K$ candidate
tokens per draft slot from OneDecode (or, on miss, from the geodesic
drafter) and submits all $K$ to the verifier in a single batched forward.
Structurally similar to Medusa-style multi-head drafting
\cite{cai2024medusa}, except candidates come from a baked flow map
rather than learned auxiliary heads.

\noindent
None of the three has the closed-form throughput model of
\Cref{sec:closedform} yet; they are listed as
\emph{implemented, not measured}.

% =================================================================
\section{Related Work}
The exactness-preserving rejection rule is from Leviathan et~al.\
\cite{leviathan2023speculative} and Chen et~al.\ \cite{chen2023speculative}.
EAGLE \cite{li2024eagle} and Medusa \cite{cai2024medusa} train auxiliary
drafter heads against the same backbone; we do not---we use a
GP-compressed copy.  This is closer to ``self-speculative''
decoding.  Production runtimes (vLLM \cite{kwon2023vllm}, TensorRT-LLM)
ship speculative paths; we are not competing on raw tok/s, we are
characterising the interaction with compression.  AttnRes is implemented
independently here.  StreamingLLM \cite{xiao2023streamingllm}'s sink-token
treatment is referenced but orthogonal.

% =================================================================
\section{Status, Limitations}
\label{sec:status}
\paragraph{Open issues.}
\begin{itemize}
\item \texttt{--ott-swarm-k 8} crashes; the SWARM-K hit count in
\Cref{tab:measured} is $0$ for that reason.
\item OneDecode/OTT-OD/OTT-SWARM acceptance rates and end-to-end tok/s
are unmeasured.
\item Tier-asymmetric speculative (8B drafter / 70B verifier) is the
deployment in which the closed-form model predicts the largest
composition win; we cannot test it on the available hardware.
\item AttnRes-on-compressed and KV-cache-projection at long context are
both implemented but unmeasured.
\end{itemize}

\paragraph{Methodological gaps.}  Single model (135M-Instruct) for
end-to-end measurement; single hardware target; no behavioural-quality
benchmarks (MMLU, GSM8K, HumanEval) on the accepted-token sequence.  The
closed-form throughput model in \Cref{sec:closedform} uses
companion-paper numbers extrapolated to this drafter; whether those
extrapolations transfer is the central uncertainty.

\section*{Acknowledgements}
The author thanks the open-source SmolLM2 release that made fast
end-to-end iteration possible and the existing speculative-decoding
literature on which the rejection rule is based.

\printbibliography
\end{document}
