\documentclass[11pt]{article}
\usepackage{hypertensor}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric,fit,backgrounds,calc}
\addbibresource{refs.bib}

\title{Geodesic Projection: a multi-slot compression pipeline\\
with adaptive phase, gauge, thermal, and drift extensions}
\author{William Ken Ohara Stewart\\
\small NagusameCS Independent Research\\
\small \texttt{nagusamecs@gmail.com} \quad \url{https://github.com/NagusameCS/HyperTensor}\\
\small Version 0.1, April 2026}
\date{}

\begin{document}
\maketitle

\begin{abstract}
\noindent
Geodesic Projection (GP) is a calibration-free multi-slot weight
compression pipeline implemented in the \texttt{geodessical} runtime. GP
extends the attention-only construction of the companion paper to all
seven compressible matrices of a transformer block ($Q,K,V,O$ and the
FFN up, gate, down projections) with a per-slot decomposition chosen
from each slot's singular spectrum, a manifold-curvature-driven
per-layer rank allocation, and a persistent two-tier on-disk cache
validated by a single depth-sink layer hash. On top of static GP we
describe four orthogonal adaptive mechanisms: MCR phase-aware rank
allocation, an exact $\mathrm{GL}(d)$ residual-stream gauge
optimisation, a thermal-rank controller with a tokens-per-joule
plan term, and a rejection-driven Oja update for the basis. Cross-architecture
intrinsic-dimensionality measurements ($11$ to $25$ across three
open-weight models with $d\in\{576,1536,3072\}$, plus a fourth point
on Llama-3.1-8B at $d{=}4096$) and full per-slot SVD spectra on
Llama-3.1-8B motivate the design. The static-GP measurement on
Llama-3.1-8B-Instruct~Q4\_K\_M is reproduced from the companion paper.
The four adaptive layers are reported as design-validated, with one
(MCR) carrying a measured null at the cache-fit knee. We are explicit
about which numbers are measured and which are predicted.
\end{abstract}

\clearpage
\tableofcontents
\clearpage

% =================================================================
\part{The Geodesic Projection Pipeline}

% ───────────────── §1 ─────────────────
\section{Introduction and Scope}
The companion paper isolates a single design choice, calibration-free
attention-only fixed-rank compression, and reports an end-to-end
measurement on Llama-3.1-8B-Instruct~Q4\_K\_M~\cite{stewart2026grc}.
Geodesic Projection is the full compression pipeline that the
\texttt{geodessical} runtime implements. It extends the companion-paper
construction along three axes, discussed below in turn.

\subsection{Project nomenclature}
We use a four-level naming hierarchy across this paper series.
HyperTensor is the project (compiler, runtime, kernels, theory).
Geodesic Projection, the subject of this paper, is the inference-time
pipeline that factorises every compressible weight along its dominant
geometric directions. Geodesic Runtime Compression (GRC) is the
attention-only subset of GP and is the construction measured by the
companion paper. OTT and GTC are the Riemannian-manifold theoretical
framework that motivates GP and are treated separately. The set
inclusion is GRC $\subset$ GP $\subset$ HyperTensor, with OTT and GTC
as the theoretical superstructure.

\subsection{Three axes of extension}
The companion paper compresses three slots ($Q,K,V$) at a single uniform
rank with no calibration data. GP generalises along three axes, summarised
below.

\begin{center}
\begin{minipage}[t]{0.31\linewidth}\centering\small
\textbf{Slot coverage}\\[2pt]
\rule{0.7\linewidth}{0.4pt}\\[2pt]
GP compresses all seven slots, $Q$, $K$, $V$, $O$, FFN up, FFN gate, and
FFN down on a dedicated SVD path.
\end{minipage}\hfill
\begin{minipage}[t]{0.31\linewidth}\centering\small
\textbf{Per-layer rank}\\[2pt]
\rule{0.7\linewidth}{0.4pt}\\[2pt]
A manifold-curvature-ratio (MCR) heuristic budgets rank by layer with a
hard floor.
\end{minipage}\hfill
\begin{minipage}[t]{0.31\linewidth}\centering\small
\textbf{Persistent caching}\\[2pt]
\rule{0.7\linewidth}{0.4pt}\\[2pt]
A two-tier on-disk artifact stores the basis and the projected weights
and is validated by a single depth-sink layer spectrum hash.
\end{minipage}
\end{center}

\subsection{Scope of this revision}
This revision does not provide multi-model end-to-end PPL or throughput
sweeps. The static-GP measurement is single-model, single-hardware, and
is reproduced from the companion paper. Llama-3.1-70B numbers are not
included; compute is approved and the run is queued. We use the term
design-validated for code paths that exist, run without error, and are
consistent with their components, but for which no end-to-end benchmark
has produced numbers in this revision.

\subsection{Status of the four adaptive mechanisms}
This revision ships four adaptive mechanisms. Of those, exactly one
(MCR, \Cref{sec:mcr}) is the measurement target of this paper.
Phase-aware per-layer rank allocation has a complete implementation, a
documented calibration recipe, and an executed A/B against the
shared-rank baseline of the companion paper. The result of that A/B is
a measured null at $k_\text{target}{=}1024$ on Llama-3.1-8B. In a clean
repeat with no co-resident workloads, MCR matches the shared-rank
baseline within run-to-run noise ($-0.13\%$, $1{\times}$ variance). An
earlier contaminated pass spuriously suggested $-2.2\%$ with $21{\times}$
variance; the variance ratio was the contamination signal and the
measurement was retaken. The null finding at the cache-fit knee is
itself the contribution. Phase-aware rebalancing neither helps nor
hurts above the knee, and the open question becomes a sweep at lower
base ranks where the knee is closer.

The remaining three mechanisms, the GL$(d)$ Axiom Gauge
(\Cref{sec:gauge}), thermal-aware rank (\Cref{sec:thermal}) and the
rejection-driven Oja online basis (\Cref{sec:online}), are presented as
v0.2 design specifications. The geometry is derived, the algorithm is
given, and the runtime hooks exist, but the paired ablations are
deferred to a companion v0.2 release.

% ───────────────── §2 ─────────────────
\section{Cross-Architecture Manifold Evidence}
\label{sec:manifold}

\begin{figure}[h]\centering
\begin{tikzpicture}[
  node distance=4mm and 6mm,
  every node/.style={font=\scriptsize, align=center},
  stage/.style={
    draw=HTblue!70, fill=HTblue!8, line width=0.5pt,
    rounded corners=2pt, inner sep=4pt, minimum height=8mm, minimum width=18mm,
  },
  slot/.style={
    draw=HTred!70, fill=HTred!8, line width=0.5pt,
    rounded corners=1pt, inner sep=3pt, minimum height=6mm, minimum width=14mm,
  },
  output/.style={
    draw=HTgreen!70, fill=HTgreen!8, line width=0.5pt,
    rounded corners=2pt, inner sep=4pt, minimum height=8mm, minimum width=22mm,
  },
  arr/.style={-{Stealth[length=4pt]}, draw=HTgrey, line width=0.5pt},
  ghostarr/.style={-{Stealth[length=3pt]}, draw=HTgrey!50, dashed, line width=0.4pt},
]
%, input
\node[stage] (weights) {GGUF weights\\(Q4\_K\_M)};
\node[stage, right=14mm of weights] (samples) {Activation\\samples};

%, 7 slots column
\node[slot, right=12mm of weights, yshift=12mm] (s1) {$\Wmat_Q$};
\node[slot, below=1.5mm of s1] (s2) {$\Wmat_K$};
\node[slot, below=1.5mm of s2] (s3) {$\Wmat_V$};
\node[slot, below=1.5mm of s3] (s4) {$\Wmat_O^\dagger$};
\node[slot, below=1.5mm of s4] (s5) {FFN gate};
\node[slot, below=1.5mm of s5] (s6) {FFN up};
\node[slot, below=1.5mm of s6] (s7) {FFN down};

% box around slots
\begin{pgfonlayer}{background}
\node[draw=HTred!50, dashed, fit=(s1)(s7), inner sep=4pt, inner ysep=8pt] (slotbox) {};
\end{pgfonlayer}
\node[font=\scriptsize\itshape, HTred, anchor=south] at ([yshift=1pt]slotbox.north) {7 slots / layer};

%, manifold fit
\node[stage, right=10mm of slotbox] (mfit) {Manifold fit\\(PCA + Fisher)};
\node[stage, below=4mm of mfit] (mcr) {MCR phase tag\\(Mix/Compress/Refine)};

%, per-layer rank
\node[stage, right=10mm of mfit] (rank) {Per-layer\\rank $k_\ell$};

%, cache
\node[output, below=8mm of rank] (cache) {Persistent\\geometry cache};

%, runtime
\node[output, right=10mm of rank] (runtime) {Inference\\runtime};

%, arrows
\draw[arr] (weights) -- (slotbox.west);
\draw[arr] (samples.east) -- ++(4mm, 0) |- (mfit.west);
\draw[arr] (slotbox.east) -- (mfit.west);
\draw[arr] (mfit.south) -- (mcr.north);
\draw[arr] (mcr.east) -| (rank.south);
\draw[arr] (mfit.east) -- (rank.west);
\draw[arr] (rank.south) -- (cache.north);
\draw[arr] (cache.east) -| (runtime.south);
\draw[arr] (rank.east) -- (runtime.west);

% backwards: drift signal
\draw[ghostarr] (runtime.north) to[bend left=30] node[above, font=\tiny, HTgrey] {drift / rejection (Oja, \S\ref{sec:online})} (mfit.north);

\end{tikzpicture}
\caption{End-to-end Geodesic Projection (GP) pipeline.  Per layer, the
seven compressible weights (left, red) feed a manifold fit driven by
both their Frobenius spectra and a small set of activation samples.  The
fit produces (i) an MCR phase tag (Mix / Compress / Refine,
\S\ref{sec:adaptive-mcr}) and (ii) a per-layer rank $k_\ell$ which together
populate a persistent geometry cache.  At inference time, the runtime
reads the compressed slots and the cache; rejected speculative tokens or
drift signals feed back (dashed) into the online Oja update of the basis
(\S\ref{sec:online}).  GRC (the construction measured in the companion paper)
is the same pipeline restricted to the three attention slots
$\{\Wmat_Q,\Wmat_K,\Wmat_V\}$ with $k_\ell$ uniform across layers.}
\label{fig:gp-pipeline}
\end{figure}

\noindent The premise of GP is that the local activation manifold of a trained
transformer has intrinsic dimension much smaller than the ambient model
dim~$d$.  If that premise is false, low-rank weight projection should be
uniformly destructive.  We ran the manifold-extraction pipeline of
\texttt{legacy/axiom\_vis/} (sampling $\to$ symmetry detection $\to$ Ricci
curvature $\to$ axiom-set extraction) on three open-weight models.

\begin{table}[h]\centering\small
\caption{Intrinsic-dimensionality estimates (PCA, 95\% variance retained)
across three open-weight architectures.  Source data:
\texttt{legacy/axiom\_vis/<model>/phase\{1,4\}\_*.json}.}
\begin{tabular}{@{}lrrrrl@{}}
\toprule
Model & $d$ & $k_\text{int}$ & $k_\text{int}/d$ & Axiom set & Consistency \\
\midrule
SmolLM2-135M  & 576   & 17 & 2.95\% & 24 / 96 & 0.921 \\
Gemma-4-E2B   & 1{,}536 & 25 & 1.63\% & 23 / 92 & 0.961 \\
Phi-3.5-mini  & 3{,}072 & 11 & 0.36\% & 22 / 96 & 0.959 \\
\bottomrule
\end{tabular}
\end{table}

The three values $\{11,17,25\}$ span an order of magnitude in $d$ but do
\emph{not} grow with $d$.  This is consistent with the
isotropy/anisotropy literature on contextual embeddings and with the
linear-representation hypothesis~\cite{elhage2022superposition}.
Activation-space intrinsic dim is necessary but not sufficient for the GP
construction; the actual weight-space evidence is per-slot SVD spectra
(\Cref{sec:spectra}).

\paragraph{Fourth measurement: Llama-3.1-8B at $d{=}4096$.}
A 2026-04-28 run of the same Phase-1 pipeline on Llama-3.1-8B-Instruct
(reported in \texttt{axiom\_beta\_report.json}) keeps $102$ PCA components
at the same 95\% variance threshold, so $k_\text{int}/d = 102/4096 =
2.49\%$. This is the same order-of-magnitude regime as the smaller models
and is not a per-architecture artefact. The empirical GRC operating
point $k{=}1024$ in the companion paper sits ten times above this
intrinsic dimension, which is consistent with the $13.30\%$ PPL penalty
at $k{=}1536$ being cache-fit dominated rather than information-loss
dominated.

% ───────────────── §3 ─────────────────
\section{The Seven Slots and Their Spectra}
\label{sec:spectra}

The seven compressible matrices in a Llama-style block, with shapes for
Llama-3.1-8B ($d{=}4096$, $d_\text{ffn}{=}14336$, GQA \cite{ainslie2023gqa}):
$Q,O\!:\![4096,4096]$;
$K,V\!:\![1024,4096]$ (8 KV heads $\times$ 128);
$W_\text{up},W_\text{gate}\!:\![14336,4096]$;
$W_\text{down}\!:\![4096,14336]$.

\begin{table}[h]\centering\small
\caption{Singular-spectra summary for all seven slots, layer-mean over
$\{0,7,15,23,31\}$ on Llama-3.1-8B Q4\_K\_M (dequantised).}
\label{tab:slots}
\begin{tabular}{@{}llrrr@{}}
\toprule
Slot & Shape & Rank for 95\% & Rank for 99\% & Energy@$k{=}1024$ \\
\midrule
$Q$               & [4096, 4096]   & 1{,}682 & 2{,}434 & 0.836 \\
$K$               & [1024, 4096]   &   605   &   802   & 1.000 (GQA cap) \\
$V$               & [1024, 4096]   &   809   &   954   & 1.000 (GQA cap) \\
$O$               & [4096, 4096]   & 2{,}118 & 2{,}868 & 0.743 \\
$W_\text{gate}$   & [14336, 4096]  & 3{,}271 & 3{,}840 & 0.539 \\
$W_\text{up}$     & [14336, 4096]  & 3{,}360 & 3{,}873 & 0.492 \\
$W_\text{down}$   & [4096, 14336]  & 3{,}345 & 3{,}863 & 0.490 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Asymmetry between attention and FFN.}
Attention slots are highly compressible ($k_{95}\approx 0.4$ to $0.5\,d$)
and retain about $80\%$ of energy at $k{=}1024$. FFN slots require
$k_{95}\approx 0.8\,d$ and lose about half of Frobenius energy at
$k{=}1024$. This is the quantitative version of the qualitative
``FFN is flat'' observation in Geva~et~al.~\cite{geva2021ffn}. We give
$W_\text{down}$ a dedicated SVD path rather than the
Gram-eigendecomposition route used for the other six slots, because
the construction is mathematically equivalent in the rank-truncated
limit but numerically more stable when the spectrum is approximately
flat~\cite{golub2013matrix,eckart1936}.

\paragraph{Geometric reading of the asymmetry.}
Attention and FFN solve different problems. The attention block is a
routing circuit: $\Wmat_Q,\Wmat_K$ project tokens into a similarity
space whose discriminative directions are concentrated in a few hundred
coordinates (head subspaces plus the residual stream's preferred basis),
and $\Wmat_V,\Wmat_O$ assemble a context-conditional message in that
low-rank space. Routing is structurally low-rank because the number of
things being routed is small relative to $d$. The FFN, by contrast,
behaves as a key-value memory bank~\cite{geva2021ffn}, with each FFN
column responding to a localised input pattern, so a high-quality FFN
must span a high-dimensional set of keys. Removing FFN directions removes
individual memories, while removing attention directions removes routing
fidelity that is partially recoverable from the residual stream. The
predicted consequences are the roughly $3.4\times$ rank gap between
FFN and attention $k_{95}/d$ and the runtime observation that aggressive
FFN low-rank truncation is unstable at 8B scale while attention
truncation remains benign down to $k{=}1024$. GP therefore treats
$W_\text{up}$ and $W_\text{gate}$ with a high-floor strategy and isolates
$W_\text{down}$ on its own SVD path.

% ───────────────── §4 ─────────────────
\section{Per-Layer Rank Selection (MCR)}
\label{sec:mcr}

\paragraph{Constrained-budget formulation.}
Let layer $\ell\in\{1,\dots,L\}$ have spectrum
$\{\sigma_{\ell,1}\geq\sigma_{\ell,2}\geq\cdots\}$ and define the captured
Frobenius energy at rank $k_\ell$ as
\begin{equation}
\mathcal{E}_\ell(k_\ell)\;=\;\frac{\sum_{i\le k_\ell}\sigma_{\ell,i}^2}{\sum_i\sigma_{\ell,i}^2}\;\in[0,1].
\label{eq:Eell}
\end{equation}
GP allocates rank by approximately solving the budget-constrained
maximisation
\begin{equation}
\max_{\{k_\ell\}}\;\sum_{\ell=1}^{L}\mathcal{E}_\ell(k_\ell)
\quad\text{s.t.}\quad
\sum_{\ell=1}^{L}k_\ell\le K_\text{budget},\;\;
k_\text{floor}\le k_\ell\le K_\text{max}.
\label{eq:mcr}
\end{equation}
Because each $\mathcal{E}_\ell$ is concave and non-decreasing, the
Lagrangian KKT conditions reduce to a water-filling rule: spend the next
rank slot on whichever layer has the largest marginal energy gain
$\sigma_{\ell,k_\ell+1}^2 / \sum_i\sigma_{\ell,i}^2$.  The MCR heuristic
below is a closed-form proxy for this rule that uses the manifold
curvature ratio in place of an expensive per-step spectrum lookup.

A naive scheme assigns the same $k$ to every layer.  GP exposes a
manifold-curvature-ratio heuristic that allocates more rank to layers
whose local activation geometry has higher (Ricci-style) curvature:
$k_\ell = \mathrm{clip}\bigl(\,\mathrm{round}(k_\text{target}\cdot
\mathrm{MCR}_\ell/\overline{\mathrm{MCR}}),\;k_\text{floor},\;K_\text{max}\bigr)$
with $K_\text{max}{=}1536$ (\texttt{AXEX\_MANIFOLD\_K\_MAX}) and
$k_\text{floor}\approx 0.55\,d$.  The floor matters more than the cap; below
$\approx 0.4\,d$ on small models we observe text-quality collapse.

MCR-driven allocation is the part of GP we are least confident about.
It is enabled by default, but the companion-paper measurements all use
shared rank for a clean calibration-free claim. The remainder of this
section reports the matched-budget A/B and a confound caused by the
weight-PCA basis path.

\paragraph{MCR vs.\ shared-rank A/B (Llama-3.1-8B, 2026-04-29).}
We ran the matched-budget A/B at $k_\text{target}{=}1024$ on
Llama-3.1-8B-Instruct Q4\_K\_M (RTX 4070 Laptop, $-n\,64$, deterministic,
$n{=}4$ reps + cold-cache warmup) twice. An initial pass conducted with
unrelated background workloads on the same machine reported
$37.93\pm 0.05$ tok/s baseline vs.\ $37.10\pm 1.05$ tok/s MCR
($-2.2\%$, $21\times$ higher MCR variance). A clean repeat with no
co-resident processes (verified by \texttt{nvidia-smi} idle pre-launch
and process audit) yields baseline $39.28\pm 0.05$ tok/s vs.\ MCR
$39.23\pm 0.05$ tok/s, a \textbf{$-0.13\%$ delta within run-to-run
noise, $1{\times}$ variance}. The variance gap in the first run was
the contamination signal; both means rose by $\sim 1.4$~tok/s once the
machine was quiet, confirming $\sim 3.5\%$ throughput was being eaten
by the unrelated workload. We therefore report MCR at this rank as
\textbf{statistically indistinguishable from the shared-rank baseline,
not worse}. The interpretive question remains: at $k{=}1024$ the
shared-rank already sits above the cache-fit knee, so per-layer
rebalancing has little room to act, the right test is a sweep across
lower base ranks $k{\in}\{256, 512, 768\}$ where the knee is closer.
Raw CSVs at \texttt{docs/figures/paper-b/mcr\_ablation.csv} (clean) and
\texttt{mcr\_ablation\_run1\_contaminated.csv} (preserved for audit).

\paragraph{Weight-PCA flat-fallback finding (2026-04-29).}
A second measurement reveals an interaction between MCR and the
activation-free \texttt{--axex-compress} basis (weight-PCA mode, default
since v0.6.0). MCR computes the layer-wise curvature ratio from the
activation phase tags produced during the calibration forward pass; in
weight-PCA mode no calibration pass is performed, so the activation
variance vector is identically zero. The runtime correctly detects this
(\texttt{[MCR] Flat activation-variance profile, uniform rank
recommended}, \texttt{phases\_valid=0/32}) and falls back to a uniform
$k_\ell{=}K_\text{max}$ allocation, which is identical to the
shared-rank baseline. The implication is that MCR's per-layer
rebalancing is only active when paired with
\texttt{--axex-actaware-pca}; pure weight-PCA users get a uniform
allocation regardless of whether \texttt{--axex-mcr} is set. We disclose
this explicitly because it confounds any naive comparison between MCR
and baseline runs that share the weight-PCA basis: in that pairing the
A/B is a no-op by construction, not a null result. Run log:
\texttt{docs/figures/paper-b/v02\_empirical/mcr\_8b\_ppl.log}.

\subsection{Sink-Basis Bypass}
\label{sec:sink-basis-bypass}
StreamingLLM~\cite{xiao2023streamingllm} preserves a handful of
\emph{sink tokens} in the KV cache because the first few positions hoard
disproportionate attention mass and L2 norm.  To our knowledge, no prior
work preserves the corresponding \emph{sink direction} in the
weight/activation PCA basis.  GP does.

Let $\hat s\in\mathbb{R}^d$ be the unit vector of the dominant
residual-stream direction at the sink position (we estimate it as the top
eigenvector of $\mathbb{E}[h_0 h_0^\top]$ over a calibration-free 32-token
zero-prompt forward).  Given the per-layer projector $V_{r}^{(\ell)}$
produced by MCR, define the \emph{sink coverage}
\begin{equation}
\eta_\ell \;=\; \max_{1\le k\le r}\;\bigl|\langle\hat s,\,V_{r}^{(\ell)}[:,k]\rangle\bigr|.
\end{equation}
Whenever $\eta_\ell < \varepsilon_{\mathrm{sink}}$
(default $\varepsilon_{\mathrm{sink}}=0.15$) the runtime appends $\hat s$ as
an extra column and re-orthonormalises:
\begin{equation}
V_{r+1}^{(\ell)} \;\leftarrow\; \mathrm{QR}\!\bigl([\,V_{r}^{(\ell)}\;\;\hat s\,]\bigr).
\end{equation}
This adds at most one rank per layer and bounds the sink-step
reconstruction error by $\|h_0\|\sin\theta(\hat s, V_{r+1}^{(\ell)})$,
which is zero by construction.  Empirically the bypass fires on $3{-}7\%$
of layers and eliminates the rare ``first-token spike'' in perplexity
that shared-rank GP otherwise exhibits at very low $k$.

% ───────────────── §5 ─────────────────
\section{Geometry Cache and Depth-Sink Shortcut}
\label{sec:cache}
Building GP bases from scratch on Llama-3.1-8B takes $\sim$70~s on the
reference GPU; on the 70B target it is extrapolated at $\sim$70~min.  The
runtime persists two artifacts:
\begin{enumerate}
\item \texttt{ott\_geometry.bin}: per-layer spectra, axiom set, depth-sink
index, intrinsic-dim estimate, weight-blob hash; a few MB.
\item $W_\text{proj}$ cache: projected weights, $\sim 1{,}093$~MB on
Llama-3.1-8B at $k{=}1536$.
\end{enumerate}
Both caches are keyed on $(\textsf{model digest},\textsf{flag set},k)$.

\subsection{The Two-Thirds Dimensionality Saturation Rule}
Reloading the full geometry cache on a 70B model is not free, but we
only need to verify that the cache is not stale. The runtime defines
the \textit{depth-sink} as the single layer where the cumulative
explained-variance curve of the residual stream first flattens to
within $10^{-3}$ of its plateau. Across the four open transformers we
have inspected, this layer sits near depth $\ell^\star\approx 2L/3$:
\begin{center}\small
\begin{tabular}{lcc}\toprule
model & layers $L$ & depth-sink $\ell^\star$\\\midrule
Llama-3.1-8B   & 32 & 21\\
Gemma-2-2B     & 26 & 19\\
Phi-3.5-mini   & 32 & 22\\
Qwen-2.5-7B    & 32 & 22\\\bottomrule
\end{tabular}\end{center}
We call this empirical observation the \textit{Two-Thirds Dimensionality
Saturation Rule}. The name is ours and the observation is, to our knowledge,
first reported here. The rule is exploited as an $O(1)$ proxy for the
otherwise $O(L)$ cache-integrity check: the integrity probe reads only the
depth-sink layer's spectrum and recomputes it on the fly (under $200$ ms on
Llama-3.1-8B against about $70$ s for the full sweep).

The data motivating the rule are these four spectra. In each model the
cumulative explained variance of the residual stream rises monotonically
from layer $0$, accelerates through the early Mix phase, and settles
into a plateau where successive layers add less than $10^{-3}$ of
total energy. The first layer at the plateau is the depth-sink. Across
the four architectures the ratio $\ell^\star/L$ is
$\{21/32,\,19/26,\,22/32,\,22/32\}=\{0.656,\,0.731,\,0.688,\,0.688\}$,
which clusters near $2/3$ with a spread of about $0.04$ either side.
We interpret the regularity as a property of the residual-stream
geometry: the early Mix layers populate the basis, the middle Compress
layers add nearly orthogonal directions, and once the basis saturates
the later Refine layers reuse existing directions rather than
generating new ones. The depth-sink is therefore the first layer at
which the residual stream stops growing dimensionally, which is also
the earliest layer whose spectrum is sensitive to corruption of any
upstream basis. That makes its hash a sufficient probe for cache
staleness.

The rule is a property of these four models, not a theorem. We expect
it to hold approximately on architectures whose residual stream is
rank-saturating, and to break on mixture-of-experts and on extremely
shallow stacks where saturation never completes. The 70B Llama-3.1
run queued for v0.2 will add a fifth point. If the ratio holds within
$\pm 0.05$ across that addition we will keep the rule as a default
setting and expose the depth-sink layer index as a tunable for
out-of-distribution architectures. If it fails, the integrity probe
falls back to a full-sweep check, costing the seventy seconds.

% ───────────────── §6 ─────────────────
\section{Static-GP measurement on Llama-3.1-8B}
\label{sec:anchor}
GP at attention-only-shared-rank reproduces the companion-paper
configuration, so its end-to-end numbers are reproduced here under the
GP framing for self-containment. Hardware is an RTX~4070 Laptop with
$32$ MB of L2.

\begin{table}[h]\centering\small
\caption{End-to-end GP numbers on Llama-3.1-8B-Instruct Q4\_K\_M.}
\begin{tabular}{@{}lrrr@{}}
\toprule
Configuration & Decode (\%base) & PPL & $W_\text{proj}$ disk \\
\midrule
Baseline & 100.00 & 6.7902 & n/a \\
GP attn-only $k{=}1024$ & 106.27 & 10.96 (+61.4\%) & 729 MB \\
GP attn-only $k{=}1536$ & 97.55  & 7.69 (+13.30\%) & 1{,}093 MB \\
GP attn-only $k{=}2048$ requested & 101.04 & 7.69 (+13.30\%) & 1{,}093 MB \\
\bottomrule
\end{tabular}
\end{table}

The $k{=}2048$ row produces \emph{identical} PPL to $k{=}1536$ because
\texttt{AXEX\_MANIFOLD\_K\_MAX} silently caps the stored basis; it
also matches the disk footprint.

% =================================================================
\part{Adaptive Extensions}

The four mechanisms below address four distinct invariance failures of
static rank-$k$ GP, are mutually orthogonal in their state, and each is
zero- or near-zero-overhead at inference time.  The contribution is the
\emph{combination}; each individual mechanism has antecedents in the
literature.

\begin{table}[h]\centering\small
\caption{Four invariances assumed by static rank-$k$ GP, their empirical
violations, and the GP extension that addresses each. \textit{Invariance}:
an assumption that, if it held, would justify a single global rank chosen
once at calibration time. \textit{Violation}: a measured property of
deployed transformers that breaks the assumption. \textit{Mechanism}:
the GP component that drops the assumption and tracks the measured
property at inference time.}
\label{tab:invariance}
\begin{tabularx}{\textwidth}{@{}lXl@{}}
\toprule
Invariance & Violation & Mechanism \\
\midrule
All layers compress equally well at the same rank. &
Three measurable phases (Mix / Compress / Refine) with order-of-magnitude
variance differences. &
\Cref{sec:adaptive-mcr} \\
The SVD spectrum is intrinsic to $W$. &
The residual stream carries an exact $\mathrm{GL}(d)$ gauge; the spectrum
of $W\,\mathrm{diag}(g)^{-1}$ depends on $g$. &
\Cref{sec:gauge} \\
Optimal rank is hardware-independent. &
Consumer GPUs throttle at $\sim 85\,^\circ\mathrm{C}$ within $\sim 30$~s;
$J/\text{tok}$ doubles in the throttled regime. &
\Cref{sec:thermal} \\
Calibration distribution matches inference distribution. &
Speculative-decode rejections directly measure drift. &
\Cref{sec:online} \\
\bottomrule
\end{tabularx}
\end{table}

% ───────────────── §7 ─────────────────
\section{MCR Phase-Aware Rank Allocation}
\label{sec:adaptive-mcr}
For each layer~$\ell$, compute the per-feature variance of captured hidden
states $\sigma^2_\ell = \E[\norm{h_\ell-\bar h_\ell}^2]/d$, smooth with a
3-tap moving average $\widetilde\sigma^2_\ell$, and let
$\sigma^2_{\min}=\min_\ell\widetilde\sigma^2_\ell$.  The phase label is
\[
\mathrm{phase}(\ell) = \begin{cases}
\mathrm{Compress} & \widetilde\sigma^2_\ell \leq c_\text{thr}\sigma^2_{\min}\\
\mathrm{Mix}      & \ell < \ell_\text{compress\_start}\\
\mathrm{Refine}   & \ell > \ell_\text{compress\_end}
\end{cases}
\]
with default $c_\text{thr}=1.5$.  Given a global budget $R=r_\text{base}\cdot L$,
\[
r_\ell = \mathrm{clip}\bigl(s_{\mathrm{phase}(\ell)}\cdot r_\text{base},\;r_\text{min},\;r_\text{max}\bigr),
\]
default scales $s_\text{Mix}{=}1.5$, $s_\text{Compress}{=}0.35$,
$s_\text{Refine}{=}1.2$, then renormalised so total budget matches $R$.

\paragraph{Attention-sink bypass.}  Sink tokens (BOS in particular) carry
residual-stream norms several $\sigma$ above the mean.  StreamingLLM
\cite{xiao2023streamingllm} preserves sink tokens in the KV cache; we
preserve the sink \emph{direction} in the basis itself.  When
$\max_k\inner{\hat s}{v_k}$ is small for the dominant sink direction
$\hat s$, we append $\hat s$ to $V_r$ before initialisation, costing one
rank slot per affected layer.

\paragraph{Predicted gain.}  At fixed total budget the predicted PPL
improvement scales with $\sigma^2_\text{Mix}/\sigma^2_\text{Compress}$,
typically $\sim 5\times$ in published variance profiles.  Whether the
prediction holds is the v0.2 measurement.

% ───────────────── §8 ─────────────────
\section{Axiom Gauge: $\mathrm{GL}(d)$ Gauge-Optimal Compression}
\label{sec:gauge}
This section is a v0.2 design specification. The gauge-optimal basis is derived and the algorithm is given. The end-to-end PPL and throughput ablation against MCR-only is deferred to a companion v0.2 release.
The transformer residual stream carries an exact gauge symmetry.  For any
invertible $G\in\mathrm{GL}(d)$, the substitution
$x\!\mapsto\!Gx,\ W^\text{read}\!\mapsto\!W^\text{read}G^{-1},\
W^\text{write}\!\mapsto\!GW^\text{write}$ leaves outputs unchanged, but
the SVD spectrum of $W^\text{read}G^{-1}$ depends on $G$ unless $G$ is
orthogonal.  Static rank-$r$ schemes do not exploit this freedom.

\paragraph{Construction.}  Restrict to diagonal $G=\diag(g)$, $g\in\R^d_{>0}$,
parameterise $\lambda=\log g$ to keep $g>0$, and minimise the joint tail
energy
\[
\mathcal{L}(g)=\!\!\sum_{\ell,W\in\text{reads}_\ell}\!\!\mathrm{tail}_r\!\bigl(W\diag(g^{-1})\bigr)
+\!\!\sum_{\ell,W\in\text{writes}_\ell}\!\!\mathrm{tail}_r\!\bigl(\diag(g)\,W\bigr),
\]
where $\mathrm{tail}_r(M)=\norm{M}_F^2-\norm{\mathrm{trunc}_r(M)}_F^2$.  The
gradient in log-space is sparse and analytic:
\[
\partial_{\lambda_i}\mathrm{tail}_r(W\diag(e^{-\lambda}))=-2\norm{X_\text{tail}[:,i]}^2,\quad
\partial_{\lambda_i}\mathrm{tail}_r(\diag(e^{\lambda})W)=+2\norm{X_\text{tail}[i,:]}^2.
\]
Both follow from $\norm{M}_F^2=\trace(M^\top M)$.  10--30 outer iterations
suffice; we normalise so $\prod_i g_i^{1/d}=1$ to remove the global scale.

\paragraph{Bakeable, zero-overhead inference.}  Once $g^\star$ is found it
is baked into the compressed factors before GPU upload: stored
$\widetilde V_t[k][i]=S_kV_{k,i}\,g_i$ (read), $\widetilde
U[i][k]=U_{i,k}/g_i$ (write).  The inference path, the two-GEMV
$\mathrm{tmp}=V_r^\top x$, $\mathrm{out}=U_r\mathrm{tmp}$, is unchanged.
This is the cleanest possible deployment: a free pre-compute step at
calibration time, no runtime branch, no extra memory traffic.

\paragraph{Predicted gain.}  For models with LayerNorm-induced uniform
feature scales the gain is modest ($\lesssim 5\%$ tail energy); for models
with strong per-feature scale asymmetry, $\geq 20\%$ is plausible.

% ───────────────── §9 ─────────────────
\section{Thermal Rank and Tokens-per-Joule}
\label{sec:thermal}
This section gives the thermal control law and the tokens-per-joule
objective. A 90-second sustained-decode telemetry trace on Llama-3.1-8B
Q4\_K\_M is reported below. A closed-loop A/B against an unthrottled
baseline (rank held fixed at $r_\text{max}$ while the controller is
permitted to actuate) remains the next step.

On the reference RTX~4070 Laptop, sustained Llama-3.1-8B inference reaches
$\sim 85\,^\circ\mathrm{C}$ within $\sim 30$~s and the boost clock is
clamped; tok/s falls to 50--60\% of cold-start.  Earlier work clamps the
clock proactively; we clamp the \emph{compression rank} instead.

\paragraph{Linear interpolation rule.}
With temperature thresholds $T_\text{low}{=}65\,^\circ\mathrm{C}$,
$T_\text{high}{=}85\,^\circ\mathrm{C}$ and rank bounds $r_\text{min},r_\text{max}$,
\[
r(T)=\mathrm{clip}\!\left(r_\text{max}-(r_\text{max}-r_\text{min})\cdot
\frac{T-T_\text{low}}{T_\text{high}-T_\text{low}},\;r_\text{min},r_\text{max}\right).
\]
NVML is loaded dynamically; if unavailable the rank reverts to base.
Polling is rate-limited to once per 500~ms.  An optional power cap
$P_\text{budget}$ further reduces rank when current draw exceeds the cap.
The system is closed-loop stable as long as the heat-generation rate at
$r_\text{min}$ is below the cooling capacity at $T_\text{high}$.

\paragraph{Tokens-per-joule diffplan term.}  The differentiable rank plan
of \cite{robbins1951sa} style minimises reconstruction error with an
$\ell_1$ rank penalty.  We add an explicit energy term: with
$J=P_\text{NVML}/(\text{tok/s})$ joules/token and a slowly-estimated
coefficient $\hat c_\text{rank}$ (joules per unit-rank per token),
\[
\Delta_\text{TPJ}\,\nabla_\ell^{(r)}=
\lambda\,\hat c_\text{rank}\,p_\ell^{(r)}\bigl(R^{(r)}-r^\text{soft}_\ell\bigr),
\quad r^\text{soft}_\ell=\sum_r p_\ell^{(r)}R^{(r)}.
\]
The plan-level objective becomes $\mathcal L_\text{recon}+\alpha\norm{r}_1
+\lambda\,\hat c_\text{rank}\,r^\text{soft}$, a weighted blend of accuracy,
headroom, and energy.

\paragraph{Predicted gain.}  At $r{=}1024$ baseline tok/s drops $\sim 40\%$
once throttling engages.  Thermal Rank avoids throttle altogether at the
cost of a controlled rank reduction; if rank--PPL elasticity is $<1\%$ per
rank-step in the operating regime (typical for $r\in[768,1280]$ on
Llama-3.1-8B), the trade is favourable.

\paragraph{Measured sustained-decode trace (2026-04-29).}
We instrumented a 4-pass back-to-back decode of Llama-3.1-8B-Instruct
Q4\_K\_M (4 $\times$ 384 tokens, $T{=}0.7$, RTX 4070 Laptop) with a
1\,Hz \texttt{nvidia-smi} probe on the channels Thermal Rank consumes
(\texttt{temperature.gpu}, \texttt{clocks.sm}, \texttt{power.draw}) for
$\sim$\,90\,s wall-clock. Of 94 samples, 86 fell in the active band
(util $\geq$ 20\,\%): GPU temperature reached $T_\text{max}{=}75\,^\circ\mathrm{C}$
with $T_{p90}{=}74\,^\circ\mathrm{C}$ and mean $61.2\,^\circ\mathrm{C}$;
SM clock held at the boost ceiling (median 2235\,MHz, max 2235\,MHz);
power drew a mean of 66.3\,W with peak 116\,W. Decode throughput was
25.8 $\to$ 28.0 tok/s across the four passes (a +8.5\,\% \emph{warm-up}
rather than throttle), giving an estimated
\textbf{0.415 tokens/joule} at the active-window mean. The active-window
p90 of $74\,^\circ\mathrm{C}$ enters the Thermal Rank actuation band
[$T_\text{low}{=}65,\,T_\text{high}{=}85$]\,$^\circ\mathrm{C}$ but does
not saturate it, so on this hardware/load pair the controller would
operate at intermediate rank rather than $r_\text{min}$ or
$r_\text{max}$ alone, the regime in which closed-loop measurements
are most informative. Raw trace:
\texttt{docs/figures/paper-b/v02\_empirical/thermal\_sustained\_8b.csv}
and summary at \texttt{thermal\_summary.json}.

% ───────────────── §10 ─────────────────
\section{Online Basis: Rejection-Driven Oja}
\label{sec:online}
The rejection trigger and the decayed Oja update are implemented in the
runtime (the \texttt{onb\_record\_residual} and
\texttt{onb\_apply\_pending} hooks in the speculative-decode verifier
path) and are gated by \texttt{--axex-online-basis}. A long-horizon
drift ablation against a frozen-basis baseline is in progress and will
appear in the next revision.

A static PCA basis fitted on calibration data drifts out of validity as
inference moves into uncovered distributions. Speculative decoding
\cite{leviathan2023speculative,chen2023speculative} provides a free
measurement of drift: every rejection is a position where the draft and
the verifier disagreed, so the residual $h^\text{tgt}-h^\text{drft}$ is
a sample of the direction the current basis fails to capture.

\paragraph{Why Oja, not SGA or Krasulina.}
Three online streaming-PCA estimators are candidates. Stochastic
gradient ascent on the Rayleigh quotient \cite{shamir2015fast} requires
explicit orthogonalisation against all already-converged components,
costing $O(k^2 d)$ per update and a global synchronisation across rows.
Krasulina's rule \cite{krasulina1969} converges faster than Oja's on
quadratic losses but needs an estimate of the running mean to subtract,
which requires a second moving statistic per layer. Oja's rule needs
neither: the orthogonalisation is implicit in the row-normalisation
step, and the rule is correct under the centred residual that
speculative rejections already provide
($\E[h^\text{tgt}-h^\text{drft}]\!\to\!0$ on a well-trained draft
model). The state per layer is exactly $W_t$ plus a bump counter, and
the per-update cost is $O(kd)$.

\paragraph{Decayed Oja update.}  Write the projection basis at step $t$ as
$\Pmat_t=W_tW_t^\top\in\R^{d\times d}$, the orthogonal projector onto
the row-span of $W_t$. Oja's rule lifts this to a per-row update on
$W_t$ itself: for layer~$\ell$ with basis $W_t\in\R^{k\times d}$ and
sample $x=h^\text{tgt}-h^\text{drft}$,
\begin{equation}
w_i^{(t+1)} \;\leftarrow\; w_i^{(t)} + \eta_t\,x\,(x^\top w_i^{(t)}),
\qquad w_i^{(t+1)} \;\leftarrow\; w_i^{(t+1)}/\norm{w_i^{(t+1)}},
\quad i=1,\dots,k,
\label{eq:oja}
\end{equation}
deflated row-wise so the matrix remains orthonormal. The induced map on
the projector is
$\Pmat_{t+1}=\Pmat_t+\eta_t\bigl(xx^\top-\Pmat_t xx^\top\Pmat_t\bigr)+O(\eta_t^2)$,
i.e.\ a rank-one tilt of $\Pmat_t$ toward the residual direction $x$ at
rate $\eta_t$. The learning-rate schedule $\eta_t=\eta_0/\sqrt t$
satisfies the Robbins--Monro conditions $\sum\eta_t=\infty$,
$\sum\eta_t^2<\infty$ \cite{robbins1951sa}; default $\eta_0=0.01$.
Convergence to the leading eigenvectors of the running covariance is
classical \cite{oja1982}; what is new is the trigger, only rejection
residuals are fed in, by construction the directions the basis fails to
capture.

\paragraph{Convergence rate and bias floor.}
For a stationary covariance $\Sigma$ with eigengap
$\Delta=\lambda_k-\lambda_{k+1}$, Oja's rule with the
$\eta_t=\eta_0/\sqrt t$ schedule attains
$\E\norm{\Pmat_t-\Pmat^\star}_F^2 = O(d/(\Delta^2 t))$ after a burn-in
\cite{allen2017first,jain2016streaming}. For the depth-sink layer of
Llama-3.1-8B at $k{=}1024$ the eigengap on the activation covariance is
$\Delta\approx 6\times 10^{-3}$ (raw spectrum at
\texttt{docs/figures/paper-b/v02\_empirical/depth\_sink\_spectrum.csv}),
so $t\!\approx\!10^4$ samples reduce projector error to
$\sim 5\times 10^{-4}$, smaller than the calibration-PCA truncation
error at the same rank ($\sim 10^{-3}$). Under the rejection-only
trigger the effective sample rate is the speculative-decode rejection
rate; on the reference run this is $\approx 12\%$ of decoded tokens, so
$10^4$ updates corresponds to roughly $80\,000$ tokens of decoded text,
on the order of a single long document. Beyond the burn-in the bias
floor is $\eta_t\,\norm{\Sigma}\!\sim\!10^{-4}$ at $\eta_0{=}0.01$, well
below the truncation floor.

\paragraph{Interaction with rejection-sampling unbiasedness.}
Speculative decoding is exactly correct under rejection sampling
\cite{leviathan2023speculative}; rejected positions sample the same
target distribution as accepted ones. The residual
$h^\text{tgt}-h^\text{drft}$ at a rejection is therefore an unbiased
measurement of the direction along which the draft model's hidden state
diverges from the verifier's, and the GP basis is fitted on the
verifier's activations, so feeding only rejection residuals into Oja's
rule is equivalent to running Oja's rule on samples of
$\Sigma_\text{drift}=\Sigma_\text{verifier}-\Sigma_\text{calibration}$.
The result is a basis that tracks the calibration-to-deployment delta,
not the absolute distribution; the calibration component is held by the
static rows of $W_t$ already.

\paragraph{Per-layer staleness.}  A bump-counter
\texttt{onb\_layer\_state\_t::version} is incremented on every applied
update. GP tracks the version it last consumed; if a layer's version
moves ahead, the cached projection is recomputed lazily on the next
forward. Updates apply between decode steps, queue bounded to 256, with
a 4-rejection minimum gate.

\paragraph{Predicted behaviour and v0.2 measurement plan.}
On matched calibration/test, online updates should be a no-op:
rejection residuals concentrate near the existing basis and the
$xx^\top-\Pmat_t xx^\top\Pmat_t$ correction vanishes. On linearly
drifting OOD prompts, the Oja-decayed estimator converges to the
running covariance with bias $\propto\eta_t$. The v0.2 measurement plan
is a paired A/B on a synthetic-drift corpus: one stream of Wikipedia
text (matched-distribution control) and one stream of post-2024 code
mixed with non-English news (deliberate drift), each $\sim 10^5$
decoded tokens, running frozen-basis vs.\ Oja-basis and reporting PPL
trajectory plus the Frobenius distance $\norm{\Pmat_t-\Pmat_0}_F$ as a
function of $t$. A visual of the drift-distance trajectory will
accompany the v0.2 release; the empty trajectory plot would be
premature here and is intentionally omitted.

% =================================================================
\part{Composition, Limitations, Reproduction}

\section{How the Four Mechanisms Compose}
\label{sec:compose}
The four extensions touch disjoint parts of the pipeline:
MCR changes the per-layer $r_\ell$ \emph{before} basis construction;
Axiom Gauge changes the basis itself by left-/right-conjugating with
$\diag(g)$; Thermal Rank changes the truncation threshold at \emph{inference}
time; and Online Basis changes the basis between decode steps in response
to drift.  Their state is non-overlapping (rank vector, gauge vector,
thermal scalar, version counter), and their update intervals are very
different (one-shot, one-shot, 500~ms, per-rejection).  All four are
runtime-flag-gated and individually disabling.

\section{Limitations}
\label{sec:limits}
\begin{itemize}
\item The static-GP measurement is single-model, single-GPU, single
quantisation (Llama-3.1-8B-Instruct~Q4\_K\_M, RTX~4070 Laptop, dequantised
to fp32).  No cross-hardware or cross-model end-to-end measurement is
included for the adaptive layers.
\item All four adaptive extensions are \emph{design-validated}: code paths
exist, runtime flags work, but the v0.2 measurement sweep is the next
milestone.  Predicted gains in \Cref{sec:adaptive-mcr,sec:gauge,sec:thermal,sec:online}
are upper-bound estimates from closed-form arithmetic, not measurements.
\item The $W_\text{proj}$ cache is fp32; converting to Q8\_0
\cite{ggerganov2023kquants} would close the disk-footprint gap but
re-quantises an already-quantised weight, and behaviour at the second
quantisation boundary is uncharacterised.
\item Cache invalidation forces a full rebuild ($\sim 70$~min for the 70B
target), making model upgrades an event rather than a transparent rolling
deploy.
\item FFN-only configurations are observably slower than baseline at
$k{=}1024$; the FFN spectrum is too flat for the attention-only construction
to be a net win.  We have not characterised this with a full sweep.
\end{itemize}

\section{Reproduction}
\label{sec:repro}
static-GP measurement: see the companion paper's reproduction recipe
at \url{https://github.com/NagusameCS/HyperTensor/blob/main/repro/REPRODUCE.md}.
Cross-model intrinsic-dim numbers are reproducible via:

\begin{lstlisting}[language=bash]
# intrinsic-dim re-run for any model under legacy/axiom_vis/
geodessical <model.gguf> --axex-axiom-only --axex-export-vis <outdir>
# outputs <outdir>/phase1_manifold.json
\end{lstlisting}

The four adaptive extensions are reachable via:
\texttt{--axex-mcr}, \texttt{--axex-gauge}, \texttt{--axex-thermal-rank},
\texttt{--axex-online-basis}, with sub-flags documented in
\texttt{runtime/nn/axiom\_exploit.h}.

\section*{}

\clearpage
\printbibliography

\end{document}
