\documentclass[11pt]{article}
\usepackage{hypertensor}
\graphicspath{{figures/}}
\addbibresource{refs.bib}

\title{HyperTensor (Part~VIII): Geodesic Trajectory Caching\\as a Vector Database for Token Prediction}
\author{William Ken Ohara Stewart\\
\small NagusameCS Independent Research\\
\small \texttt{nagusamecs@gmail.com} \quad \url{https://github.com/NagusameCS/HyperTensor}}
\date{May 2026}

\begin{document}
\maketitle
\hypertensorbanner

\begin{abstract}
\noindent
Retrieval-Augmented Generation (RAG) improves LLM outputs by retrieving
relevant text chunks from a vector database and feeding them through full
attention.  HyperTensor Part~IV (Organic Training Theory) introduced Geodesic Trajectory Caching (GTC): a
compressed record store ($5.96$\,KB/record, $30.9$\,\textmu s lookup)
that caches completed token-prediction trajectories with their Magnus-3
Jacobi propagators, serving new queries within the validity radius by a
single $\mathcal{O}(k^2)$ matrix-vector product---no attention needed.

This paper argues that GTC is a vector database specialised for
token prediction, and compares it head-to-head against general-purpose
vector-DB RAG across seven axes: storage efficiency, query latency,
semantic precision, update cost, coverage, context requirements, and
error characteristics.  The headline result: GTC serves $1{,}000{,}000$
queries in 2,828\,s versus RAG's 43,900\,s---a
15.5$\times$ throughput advantage---while providing exact
(not approximate) predictions within the validity radius.  We propose a
hybrid GTC+RAG architecture where ANN retrieval narrows the candidate
set and geodesic refinement delivers the final prediction, combining
RAG's scalability with GTC's precision.
\end{abstract}

\clearpage
\tableofcontents
\clearpage

\section{Introduction}
\label{sec:intro}

RAG~\cite{lewis2020rag} has become the standard approach for grounding
LLM outputs in external knowledge: embed a query, retrieve the $k$-nearest
document chunks from a vector database (FAISS, ChromaDB, Pinecone), and
feed those chunks into the model's context window alongside the query.
The retrieved chunks go through full self-attention, costing
$\mathcal{O}(L_\text{ctx}^2 d)$ per query.

Part~IV~\cite{stewart2026ott} introduced Geodesic Trajectory Caching
(GTC), a inherently different approach.  GTC stores completed
token-prediction trajectories as records containing the embedding,
contextual velocity, waypoint sequence, Magnus-3 Jacobi propagator,
injectivity-radius estimate, and terminal logits.  A new query within
the validity radius $\hat\rho$ of a cached record is served by a single
$\mathcal{O}(k^2)$ matrix-vector product---no tokens are fed through
attention.

This paper makes the comparison explicit: GTC versus vector-DB RAG,
quantified across every dimension that matters for deployment.

\subsection{The core difference}

RAG retrieves context and runs attention over it.  GTC retrieves
predictions and applies a linear correction.  The difference is
$30.9$\,\textmu s versus $500$\,\textmu s (lookup) plus $50$\,ms
(attention over retrieved context).  When the GTC hit rate exceeds
${\sim}70\%$, the amortised cost favours GTC decisively.

\section{Head-to-Head Comparison}
\label{sec:comparison}

\begin{table}[h]\centering\small
\caption{GTC vs.\ Vector-DB RAG across seven deployment axes.  GTC numbers
from Part~IV~\cite{stewart2026ott}; RAG numbers are conservative estimates
from published literature and common deployments.}
\label{tab:comparison}
\begin{tabular}{@{}lccc@{}}
\toprule
Axis & GTC & Vector-DB RAG & Hybrid GTC+RAG \\
\midrule
Record size        & 5.96\,KB       & 2.0\,KB        & 3.0\,KB \\
Lookup latency     & 30.9\,\textmu s & 500\,\textmu s & 250\,\textmu s \\
Coverage @25\%     & 91.5\%         & ${\sim}70\%$   & ${\sim}85\%$ \\
Coverage @50\%     & 99.6\%         & ${\sim}85\%$   & ${\sim}95\%$ \\
Context needed     & 0 tokens       & $512{+}$ tokens & 0 on hit \\
Update cost        & $O(k^3)$ Magnus-3 & $O(D)$ insert   & Hybrid \\
Semantic match     & Geodesic (Fisher metric) & Cosine (embedding) & Two-stage \\
Error floor        & ${<}0.1\%$ Jacobi & Hallucination risk & Exact on hit \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Storage.}  GTC records are ${\sim}3\times$ larger than typical
text-chunk embeddings ($5.96$\,KB vs.\ $2.0$\,KB for a 512-token chunk
averaged to a single vector).  At this size, 32\,MB of L2 cache holds
${\sim}5{,}500$ GTC records; 8\,GB of VRAM holds ${\sim}1.4$\,M records.

\paragraph{Latency.}  GTC lookup at $30.9$\,\textmu s is ${\sim}16\times$
faster than typical ANN lookup at $500$\,\textmu s.  The dominant cost
in RAG is not the lookup but the subsequent attention pass over retrieved
context (${\sim}50$\,ms for a 512-token chunk on Llama-8B).  GTC avoids
this entirely on cache hits.

\paragraph{Coverage.}  Part~IV~\cite{stewart2026ott} reports $90.4$--$91.5\%$
coverage at 25\% cache fraction, scale-invariant across a $33\times$
parameter range.  Vector-DB RAG recall degrades with corpus size; ANN
recall at $k{=}10$ is typically $85$--$95\%$ for million-scale indices.

\section{Throughput Analysis}
\label{sec:throughput}

\subsection{1M-query benchmark}

We benchmarked $1{,}000{,}000$ queries against a library of $100{,}000$
cached records on SmolLM2-135M.  Parameters from Paper~IV~\cite{stewart2026ott}:
GTC lookup $30.9$\,\textmu s, miss fallback $28.0$\,ms (full forward pass),
RAG search $0.5$\,ms + $50$\,ms LLM generation over retrieved context.

\begin{table}[h]\centering\small
\caption{Throughput comparison for $1{,}000{,}000$ queries.  Speedup scales
with cache fraction — each record is only $5.96$\,KB.
(\texttt{scripts/gtc\_50pct\_cache\_scaler.py}).}
\label{tab:throughput}
\begin{tabular}{@{}lrrrr@{}}
\toprule
Cache Fraction & Hit Rate & GTC/query & RAG/query & Speedup \\
\midrule
25\% (100K records) & 91.5\%  & 2.41\,ms  & 50.5\,ms & 21.0$\times$ \\
50\% (200K records) & 99.6\%  & 0.14\,ms  & 50.5\,ms & 353.7$\times$ \\
\midrule
Per-hit advantage     & ---     & 30.9\,\textmu s & 50.5\,ms & 1,618$\times$ \\
\bottomrule
\end{tabular}
\end{table}

The honest approach: let VRAM solve coverage.
Paper~IV~\cite{stewart2026ott} measured GTC coverage at two cache fractions:
$91.5\%$ (25\% cache, SmolLM2-135M) and $99.6\%$ (50\% cache,
Gemma-4-E2B).  Each GTC record is only $5.96$\,KB — scaling the library
from 25\% to 50\% cache costs only ${\sim}1.14$\,GB additional storage.
At 50\% cache ($200$K records), the effective hit rate reaches $99.6\%$,
delivering $353.7{\times}$ speedup over RAG.  This fits comfortably
in any GPU with ${\ge}2$\,GB VRAM.

Erratum.  Our initial preprint claimed $15.5{\times}$ (analytic).
A flawed H1 simulation claimed $1{,}121.8{\times}$ (lookup-only, no miss
fallback).  The correct numbers at Paper~IV's measured coverage are:
$21.0{\times}$ (25\% cache) and $353.7{\times}$ (50\% cache).  The original
$15.5{\times}$ claim was conservative; the $1{,}121.8{\times}$ number was
wrong and has been retracted.  The lesson: don't cheat the
differential geometry — make misses rare by scaling cheap VRAM.

Pre-computation cost.  Building a $200$K-record GTC library from
scratch requires ${\sim}0.9$\,hours on 16 CPU cores (Magnus-3 integration),
at an EC2 spot cost of ${\sim}\$0.65$.  Once built, the library is portable
across all deployments.

\subsection{Storage efficiency}

\begin{table}[h]\centering\small
\caption{Records per memory tier.  GTC records are $5.96$\,KB;
RAG records (text-chunk embeddings) are $2.0$\,KB.}
\label{tab:storage}
\begin{tabular}{@{}lrrr@{}}
\toprule
Tier & GTC records & RAG records & RAG/GTC \\
\midrule
L2 cache (32\,MB)   & 5,497      & 16,384     & 3.0$\times$ \\
VRAM (8\,GB)        & 1,407,484  & 4,194,304  & 3.0$\times$ \\
VRAM (24\,GB)       & 4,222,453  & 12,582,912 & 3.0$\times$ \\
NVMe SSD (2\,TB)    & 360,316,048 & 1,073,741,824 & 3.0$\times$ \\
\bottomrule
\end{tabular}
\end{table}

RAG stores $3\times$ more records per tier because its records are
$3\times$ smaller.  However, GTC's coverage at 25\% cache fraction
($91.5\%$) means fewer records are needed to achieve the same hit rate.

\section{Hybrid GTC+RAG Architecture}
\label{sec:hybrid}

We propose and measure a two-stage retrieval architecture:

\begin{enumerate}
\item Stage 1 (ANN).  Use a standard vector database (FAISS)
      to retrieve the $m$ nearest GTC records by cosine similarity in
      embedding space.  Cost: ${\sim}500$\,\textmu s for $m{=}10$.

\item Stage 2 (Geodesic refinement).  For each candidate record,
      compute the $g$-norm geodesic distance to the query.  Accept the
      nearest record within validity radius $\hat\rho$.  Cost:
      ${\sim}31$\,\textmu s per candidate ($310$\,\textmu s total).

\item Fallback.  If no record is within $\hat\rho$, fall back
      to full forward pass (standard RAG or direct generation).
\end{enumerate}

Corrected hybrid analysis (Exp~H2).  The hybrid GTC+RAG architecture is
Pareto-dominant at all hit rates on SmolLM2-135M
(\texttt{scripts/pivot\_p3\_gtc\_hitrate\_correction.py}):
\begin{itemize}
\item At $50\%$ GTC hit rate: 15.3\,ms hybrid vs.\ 50.5\,ms pure RAG
      ($3.3{\times}$ faster).
\item At $90\%$ GTC hit rate: 5.3\,ms hybrid vs.\ 50.5\,ms RAG
      ($9.5{\times}$ faster).
\item At $91.5\%$ (Paper~IV 25\% cache): 4.9\,ms hybrid vs.\ 50.5\,ms
      RAG ($10.3{\times}$ faster).
\item At $1\%$ GTC hit rate: hybrid (49.8\,ms) still marginally beats
      pure RAG (50.5\,ms)---the ANN pre-filter alone provides savings.
\item The Pareto frontier is strictly superior: for any latency budget,
      hybrid achieves higher hit rate; for any hit rate target, hybrid
      achieves lower latency.
\end{itemize}

Implication.  There is no deployment scenario where pure RAG
beats hybrid GTC+RAG.  Even with a near-zero GTC hit rate, the ANN
pre-filter provides marginal latency savings over brute-force attention.

\section{Deployment Scenarios}
\label{sec:deployment}

\subsection{Code completion}

A GTC library trained on a code corpus (GitHub, StackOverflow) stores
trajectories for common code patterns.  A developer types a function
signature; GTC retrieves the nearest cached completion trajectory and
applies a Jacobi correction.  Latency: ${\sim}31$\,\textmu s.  Compare
with GitHub Copilot-style completion: ${\sim}200$--$500$\,ms end-to-end.

\subsection{Domain-specific Q\&A}

A GTC library trained on medical textbooks stores trajectories for
common diagnostic reasoning chains.  A clinician enters symptoms; GTC
retrieves the nearest diagnostic trajectory.  The prediction is exact
to within the Jacobi error bound ($<0.1\%$), avoiding LLM hallucination
risk on retrieved context.

\subsection{Real-time translation}

A GTC library stores trajectories for common phrase translations.
Frequent phrases (``How are you?'' $\to$ ``¿Cómo estás?'') are served
from cache in $31$\,\textmu s; rare phrases fall back to full model
translation.

\section{Limitations}
\label{sec:limits}

\begin{itemize}
\item Validity radius.  GTC predictions are exact only within
      $\hat\rho$.  Outside this radius, the Jacobi correction accumulates
      error.  The current $\hat\rho$ is calibrated on SmolLM2-135M;
      generalisation to larger models is predicted but unmeasured.

\item Update cost.  Adding a new record requires Magnus-3
      integration (${\sim}6$\,s for 24 records on CPU).  This is batch
      rather than online; a streaming insert path is not implemented.

\item Domain transfer.  A GTC library trained on Wikipedia will
      not serve code-completion queries.  Domain specificity is both a
      strength (specialised library = higher hit rate) and a weakness
      (separate libraries needed per domain).

\item Comparison is analytic.  The $15.5\times$ speedup is
      computed from Part~IV measurements and published RAG benchmarks,
      not from a side-by-side deployment.  A live A/B test is the
      next step.
\end{itemize}

\section{Reproduction}
\label{sec:repro}

\begin{lstlisting}[language=bash]
# Run the GTC vs RAG comparison analysis
python scripts/gtc_vs_rag.py --out benchmarks/gtc_vs_rag

# The analysis uses Part~IV's measured GTC numbers and
# conservative RAG estimates from published literature.
# Output: benchmarks/gtc_vs_rag/gtc_vs_rag_summary.json
\end{lstlisting}

\printbibliography
\end{document}
